---
title: "1. Basic Usage"
author: "Nikhil Kaza"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{1. Basic Usage}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

```{r load}
#| code-fold: true
library(shapeindices)
library(sf)
library(ggplot2)
library(dplyr)

theme_set(theme_minimal(base_size = 11))
```

This vignette is a practical, code-first tour of the package `shapeindices`. The goal of the package is to provide indices for `sf`, `sfc` objects that describe the shape/mass configurations. In this package, I demonstrate some key indices and approaches. For deep dive into specific index, see the corresponding vignettes.

# Three basic shapes

Three small synthetic polygons, built with base `sf` calls, cover a
useful range of boundary behaviour: a **star** (deep, evenly-spaced
notches), a **spiral** (a single winding corridor, no notches at all),
and a **blob with a hole** (an irregular ring with an interior void).

```{r shapes}
#| code-fold: true
make_star <- function(n_points, r_outer = 1, r_inner = 0.5, center = c(0, 0)) {
  n <- n_points * 2
  angles <- seq(pi / 2, pi / 2 + 2 * pi, length.out = n + 1)[1:n]
  radii  <- rep(c(r_outer, r_inner), n_points)
  x <- center[1] + radii * cos(angles)
  y <- center[2] + radii * sin(angles)
  coords <- rbind(cbind(x, y), c(x[1], y[1]))
  st_polygon(list(coords))
}

make_turtle_path <- function(n_steps, angle_deg, step0 = 1, step_growth = 0) {
  angle <- 0
  pos   <- c(0, 0)
  pts   <- matrix(pos, ncol = 2)
  step  <- step0
  for (i in seq_len(n_steps)) {
    angle <- angle + angle_deg * pi / 180
    pos   <- pos + step * c(cos(angle), sin(angle))
    pts   <- rbind(pts, pos)
    step  <- step + step_growth
  }
  st_linestring(pts)
}

make_spiral <- function(n_steps = 48, angle_deg = 24, step0 = 0.35,
                         step_growth = 0.09, width = 0.35) {
  st_buffer(make_turtle_path(n_steps, angle_deg, step0, step_growth),
            dist = width, endCapStyle = "FLAT", joinStyle = "MITRE", mitreLimit = 3)
}

make_blob_hole <- function(n = 14, seed = 1, roughness = 0.55, hole_frac = 0.3) {
  set.seed(seed)
  angles <- sort(runif(n, 0, 2 * pi))
  radii  <- 1 + roughness * (runif(n) - 0.5) * 2
  outer  <- cbind(radii * cos(angles), radii * sin(angles))
  outer  <- rbind(outer, outer[1, ])

  # the hole is the SAME irregular outline scaled down by sqrt(hole_frac)
  # (so hole area / outer area ~= hole_frac) and traced in reverse, so sf
  # treats it as an interior ring rather than a second shell. Scaling every
  # radius by the same constant keeps the hole strictly inside the outer
  # boundary at every angle, so the result is always a valid polygon.
  hole <- outer[nrow(outer):1, ] * sqrt(hole_frac)

  st_polygon(list(outer, hole))
}

star      <- make_star(6, r_outer = 1, r_inner = 0.4)
spiral    <- make_spiral()
blob_hole <- make_blob_hole()

basic_shapes <- list(star = star, spiral = spiral, "blob with hole" = blob_hole)

# geom_sf() + facet_wrap() can't use free scales, so normalise each shape
# onto a common bounding box (center it, scale to unit span) before faceting
normalize_geom <- function(geom, center, scale) (geom - center) * scale
shapes_norm <- lapply(names(basic_shapes), function(nm) {
  g      <- st_sfc(basic_shapes[[nm]])
  bb     <- st_bbox(g)
  center <- unname(c((bb["xmin"] + bb["xmax"]) / 2, (bb["ymin"] + bb["ymax"]) / 2))
  span   <- max(bb["xmax"] - bb["xmin"], bb["ymax"] - bb["ymin"])
  st_sf(shape = nm, geometry = normalize_geom(g, center, 1 / span))
})
shapes_sf <- do.call(rbind, shapes_norm)

ggplot(shapes_sf) +
  geom_sf(fill = "steelblue", alpha = 0.6, color = "grey20") +
  facet_wrap(~ shape) +
  theme_void(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```


# The thirteen indices

Each index has its own function `*_index()` that takes, sf or sfc objects as input.  The following table briefly describes the conceptual framing of the index and shows, as a thumbnail, a shape/mass distribution that achieves close to the maximum value. These are just some examples and the emphasis is on showing some unexpected profiles that score high on these indices. While it is tempting to think that there is an unique shape and a mass profile configuration (e.g. center-weighted disk) that maximises the index, the following table shows the folly of such naivete.

```{r shape-thumb-helper, echo=FALSE}
# a small base64-embedded PNG of a shape's own outline, embedded directly
# in the table below and reused later as a column header
shape_thumb <- function(geom, size_px = 60, col = "grey75") {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(st_geometry(st_sfc(geom)), col = col, border = "grey30")
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/><br>', uri, size_px, size_px)
}
# index_fn, if supplied, appends that index's own actual value on this
# exact shape to the label, so the table shows the claim, not just the
# picture
thumb_cell <- function(geom, label, index_fn = NULL, col = "grey75", size_px = 60, ...) {
  lbl <- if (is.null(index_fn)) label else sprintf("%s (%.2f)", label, index_fn(geom, ...)$index)
  paste0(shape_thumb(geom, size_px = size_px, col = col), lbl)
}

# a weighted counterpart: colours each triangle of a FINE mesh by its own
# weight (coarse per-triangle weight can't resolve a smooth density field -
# a raw CDT triangle of a disk already spans its full radius, so a single
# weight per triangle can't "concentrate" anything - subdivide_mesh() first
# so nearby triangles can actually carry different weights). index_fn, if
# supplied, is called on that SAME fine mesh/weight - the number shown is
# exactly what the picture depicts, not a separate, coarser calculation
weighted_thumb <- function(geom, weight_fn, label, index_fn = NULL, size_px = 60,
                            palette = "YlOrRd", max_depth = 4, ...) {
  tri <- subdivide_mesh(geom, max_depth = max_depth)
  ctr <- st_coordinates(st_centroid(st_union(st_sfc(geom))))
  tc  <- st_coordinates(st_centroid(tri))
  w   <- weight_fn(tc, ctr)
  w_scaled <- (w - min(w)) / (max(w) - min(w) + 1e-9)
  cols <- hcl.colors(20, palette, rev = TRUE)[pmax(1, ceiling(w_scaled * 20))]
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(st_geometry(tri), col = cols, border = NA)
  plot(st_geometry(st_sfc(geom)), border = "grey20", lwd = 0.6, add = TRUE)
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  lbl <- label
  if (!is.null(index_fn)) {
    prep <- list(poly = st_geometry(st_sfc(geom)), tri = tri)
    lbl <- sprintf("%s (%.3f)", label, index_fn(geom, prep = prep, weight = w, ...)$index)
  }
  paste0(sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/><br>', uri, size_px, size_px), lbl)
}
# a centre-heavy field (also symmetry-preserving, since it's a function of
# radius alone) - satisfies every weighted example below except convexity's.
# The tiny offset keeps every triangle's own weight strictly positive - a
# handful of index functions divide by weight-derived quantities internally
# and a hard zero at the shape's own edge can hit that
radial_field <- function(tc, ctr, power = 2) {
  d <- sqrt((tc[, 1] - ctr[1])^2 + (tc[, 2] - ctr[2])^2)
  (1 - d / max(d))^power + 0.01
}

edge_field <- function(tc, ctr, power = 2) {
  d <- sqrt((tc[, 1] - ctr[1])^2 + (tc[, 2] - ctr[2])^2)
  (d / max(d))^power + 0.01
}

# convexity_index()'s own weighted form tolerates ANY weighting, not just a
# symmetric one, as long as the shape itself is convex - a plain left-to-right gradient makes that point visually
linear_field <- function(tc, ctr) (tc[, 1] - min(tc[, 1])) / (max(tc[, 1]) - min(tc[, 1])) + 0.01
# for the dumbbell specifically: centre-heavy around BOTH bell ends
# (hardcoded at (0,0) and (0,5), matching make_dumbbell()'s own defaults)
# rather than the shape's own overall centroid, which sits in the empty
# middle of the connecting bar - weight concentrated where the mass
# actually is, a clearer illustration than weighting the bar itself.
# Symmetric across both ends by construction, so directional_balance_index()
# still reads exactly 1 (verified) - moment_isotropy_index() does NOT: no
# weighting of a 2-fold-symmetric shape can reach 1 there (see
# vignette("g-understanding-moment-isotropy-index")'s own "at least
# 3-fold symmetry" finding), which is why the pinwheel, not the dumbbell,
# illustrates that row instead.
dumbbell_field <- function(tc, ctr, power = 2) {
  d1 <- sqrt((tc[, 1] - 0)^2 + (tc[, 2] - 0)^2)
  d2 <- sqrt((tc[, 1] - 0)^2 + (tc[, 2] - 5)^2)
  d_near <- pmin(d1, d2)
  (1 - d_near / max(d_near))^power + 0.01
}
```


```{r index-table, echo=FALSE}
make_square <- function(half = 1) st_polygon(list(rbind(
  c(-half, -half), c(half, -half), c(half, half), c(-half, half), c(-half, -half))))
make_rectangle <- function(w, h) st_polygon(list(rbind(
  c(0, 0), c(w, 0), c(w, h), c(0, h), c(0, 0))))
make_regular_ngon <- function(n, r = 1) {
  angles <- seq(pi / 2, pi / 2 + 2 * pi, length.out = n + 1)[1:n]
  coords <- cbind(r * cos(angles), r * sin(angles))
  st_polygon(list(rbind(coords, coords[1, ])))
}
# a 1% radial jitter, not a true st_buffer() circle - a perfectly regular
# circle's boundary points sit simultaneously near-cocircular by
# construction, which makes RTriangle's CDT triangulation pick a
# near-degenerate diagonal at many points at once; subdivide_mesh()'s own
# recursive refinement of that mesh is then sensitive to machine-epsilon
# input differences (confirmed: platform-level floating-point noise alone
# flipped enough of those ties to move a weighted, subdivided index by
# over a percentage point - see the footnote on the table below). The
# jitter is visually indistinguishable from a circle (well under 1% of
# radius) but breaks the exact symmetry, so no nearby diagonal choice is
# ever a genuine tie - verified stable under further perturbation down to
# 1e-9, unlike the true circle it replaces.
make_disk <- function(r = 1, n = 96, jitter = 0.01, seed = 1) {
  set.seed(seed)
  angles <- seq(0, 2 * pi, length.out = n + 1)[1:n]
  radii  <- r * (1 + jitter * (runif(n) - 0.5))
  coords <- cbind(radii * cos(angles), radii * sin(angles))
  st_polygon(list(rbind(coords, coords[1, ])))
}
make_star <- function(n_points, r_outer, r_inner) {
  n <- n_points * 2
  angles <- seq(pi / 2, pi / 2 + 2 * pi, length.out = n + 1)[1:n]
  radii  <- rep(c(r_outer, r_inner), n_points)
  x <- radii * cos(angles); y <- radii * sin(angles)
  st_polygon(list(rbind(cbind(x, y), c(x[1], y[1]))))
}
make_pinwheel <- function(n_arms, arm_len = 5, arm_width = 0.7, hub_r = 0.3) {
  angles <- seq(0, 2 * pi, length.out = n_arms + 1)[1:n_arms]
  hub  <- st_buffer(st_sfc(st_point(c(0, 0))), dist = hub_r, nQuadSegs = 20)[[1]]
  arms <- lapply(angles, function(a) {
    rot  <- matrix(c(cos(a), sin(a), -sin(a), cos(a)), 2, 2)
    rect <- st_polygon(list(rbind(c(0, -arm_width / 2), c(arm_len, -arm_width / 2),
                                   c(arm_len, arm_width / 2), c(0, arm_width / 2), c(0, -arm_width / 2))))
    rect * rot
  })
  Reduce(function(a, b) st_union(st_sfc(a))[[1]] |> st_union(st_sfc(b)[[1]]), arms, hub)
}
make_dumbbell <- function(r, cy = 5) {
  top <- st_buffer(st_sfc(st_point(c(0, cy))), dist = r, nQuadSegs = 40)[[1]]
  bot <- st_buffer(st_sfc(st_point(c(0, 0))), dist = r, nQuadSegs = 40)[[1]]
  bar <- st_polygon(list(rbind(c(-0.15, 0), c(0.15, 0), c(0.15, cy), c(-0.15, cy), c(-0.15, 0))))
  st_union(st_sfc(top, bot, bar))[[1]]
}
make_flower <- function(n_petals = 6, r_out = 1, r_in = 0.55, n_pts = 240) {
  ang <- seq(0, 2 * pi, length.out = n_pts + 1)[1:n_pts]
  r <- r_in + (r_out - r_in) * (0.5 + 0.5 * cos(n_petals * ang))
  coords <- cbind(r * cos(ang), r * sin(ang))
  st_polygon(list(rbind(coords, coords[1, ])))
}
# width_length_ratio_index()'s own three-way blind spot, all still == 1:
# a hole (make_square_with_hole(), same as vignette j), needle-thin spikes
# (make_star() with a tiny inner radius, same as vignette j's own "Spikes"
# section), and disconnected pieces (new here - none of the multi-part
# shapes elsewhere in this package happen to have a square envelope, since
# they're built to demonstrate a DIFFERENT pathology, dispersal, instead)
make_square_with_hole <- function(outer_half = 1, hole_frac = 0.5) {
  outer <- rbind(c(-outer_half, -outer_half), c(outer_half, -outer_half),
                 c(outer_half, outer_half), c(-outer_half, outer_half), c(-outer_half, -outer_half))
  hh <- outer_half * sqrt(hole_frac)
  hole <- rbind(c(-hh, -hh), c(-hh, hh), c(hh, hh), c(hh, -hh), c(-hh, -hh))
  st_polygon(list(outer, hole))
}
mk_corner_sq <- function(x0, y0, s) st_polygon(list(rbind(
  c(x0, y0), c(x0 + s, y0), c(x0 + s, y0 + s), c(x0, y0 + s), c(x0, y0))))
make_scattered_corners <- function() st_union(st_sfc(
  mk_corner_sq(0, 0, 1.3), mk_corner_sq(2.7, 0, 1.3),
  mk_corner_sq(2.6, 2.6, 1.4), mk_corner_sq(0, 3.4, 0.6)))

ex_rect     <- make_rectangle(2, 1)
ex_octagon  <- make_regular_ngon(8)
ex_hexagon  <- make_regular_ngon(6)
ex_square   <- make_square()
ex_disk     <- make_disk()
ex_pinwheel <- make_pinwheel(4)
ex_dumbbell <- make_dumbbell(1)
ex_flower    <- make_flower()
ex_star_mild <- make_star(6, 1, 0.65)
ex_hole      <- make_square_with_hole()
ex_spike     <- make_star(8, 1, 0.02)
ex_parts     <- make_scattered_corners()

warn_col <- "#e2725b"

idx_table <- data.frame(
  fn = c("convexity_index()", "moment_of_inertia_index()", "moment_isotropy_index()",
         "directional_balance_index()", "span_index()", "radial_concentration_index()",
         "depth_index()", "hull_ratio_index()", "polsby_popper_index()",
         "width_length_ratio_index()", "reock_index()", "detour_index()", "exchange_index()"),
  description = c(
    "Fraction of a connecting line between two mass-weighted random points that falls outside the shape",
    "Actual polar moment of inertia about the mass centroid vs. the same total mass optimally (radially) arranged",
    "Ratio of the mass distribution's own principal moments (elongation only, ignoring overall compactness)",
    "First-harmonic bearing bias of the mass distribution about its own centroid",
    "Mean distance between two mass-weighted random points vs. the same mass optimally arranged",
    "Mean distance from a mass-weighted random point to the shape's own geometric median, vs. optimal",
    "Mean distance from a mass-weighted random point to the boundary, vs. optimal",
    "polygon area / convex hull area", "4*pi*area / perimeter^2",
    "minimum bounding rectangle: shorter side / longer side", "area / minimum bounding circle area",
    "equal-area-circle perimeter / convex hull perimeter",
    "shape's own area shared with an equal-area circle at its centroid"
  ),
  ex_weighted = c(
    weighted_thumb(ex_rect, linear_field, "rectangle, left-to-right gradient", index_fn = convexity_index, n_quad = 1),
    weighted_thumb(ex_hexagon, radial_field, "hexagon, centre-weighted", index_fn = moment_of_inertia_index),
    weighted_thumb(ex_pinwheel, radial_field, "pinwheel, centre-weighted", index_fn = moment_isotropy_index),
    weighted_thumb(ex_disk, edge_field, "disk, edge-weigthed", index_fn = directional_balance_index),
    weighted_thumb(ex_square, radial_field, "square, centre-weighted", index_fn = span_index, n_quad = 1),
    weighted_thumb(ex_flower, radial_field, "flower, centre-weighted", index_fn = radial_concentration_index),
    weighted_thumb(ex_disk, radial_field, "disk, centre-weighted", index_fn = depth_index),
    "No weighted form", "No weighted form", "No weighted form",
    "No weighted form", "No weighted form", "No weighted form"
  ),
  ex_unweighted = c(
    thumb_cell(ex_rect, "rectangle", index_fn = convexity_index),
    thumb_cell(ex_hexagon, "hexagon", index_fn = moment_of_inertia_index),
    thumb_cell(ex_pinwheel, "pinwheel", index_fn = moment_isotropy_index),
    thumb_cell(ex_dumbbell, "dumbbell", index_fn = directional_balance_index),
    thumb_cell(ex_star_mild, "star", index_fn = span_index),
    thumb_cell(ex_octagon, "octagon", index_fn = radial_concentration_index),
    thumb_cell(ex_disk, "disk", index_fn = depth_index),
    thumb_cell(ex_square, "square", index_fn = hull_ratio_index),
    thumb_cell(ex_disk, "disk", index_fn = polsby_popper_index),
    paste(
      thumb_cell(ex_hole, "hole punched through it", index_fn = width_length_ratio_index, col = warn_col, size_px = 48),
      thumb_cell(ex_spike, "needle-thin spikes", index_fn = width_length_ratio_index, col = warn_col, size_px = 48),
      thumb_cell(ex_parts, "3 disconnected pieces", index_fn = width_length_ratio_index, col = warn_col, size_px = 48),
      sep = "<br>"
    ),
    thumb_cell(ex_disk, "disk", index_fn = reock_index),
    thumb_cell(ex_octagon, "octagon", index_fn = detour_index),
    thumb_cell(ex_star_mild, "star", index_fn = exchange_index)
  )
)
knitr::kable(format = "html", idx_table, escape = FALSE,
             col.names = c("Function", "Description", "Ex: weighted ~ 1", "Ex: unweighted ~ 1"))
```

The three thumbnails on `width_length_ratio_index`'s own row aren't a
*good* example, but a caveat. A hole punched clean through a
shape, needle-thin spikes, guitar picks, dendritic shapes with square convex hull and a shape split into several pieces all still score close to `1` here, because
the index only ever looks at the minimum-area
bounding rectangle.  See
`vignette("j-understanding-classical-indices", package = "shapeindices")`
for the other blind spots this family of indices shares.



Called directly on a shape, for example,:

```{r indices-direct}
convexity_index(star)$index
moment_of_inertia_index(spiral)$index
hull_ratio_index(blob_hole)$index
```

Running the seven mesh-based indices separately re-triangulates the
polygon up to seven times. `shape_indices()` triangulates once and
returns all thirteen as a named vector, which might have implications for speed.

```{r indices-combined}
round(sapply(basic_shapes, shape_indices), digits = 3)
```

The spiral scores low on every index (a long, narrow, winding
corridor is far from convex and far from disk-like); the blob's hole
drags its indices down relative to a solid blob of the same outer
boundary; the star sits in between, close to its `hull_ratio_index` twin but
lower on `convexity_index`, which - unlike `hull_ratio_index` - is sensitive
to the *evenly-spaced* notches cut into it.

`polsby_popper_index` in particular penalises the blob's hole twice over, not
once: `area` is already net of the hole (every index here measures net
area), but `perimeter` grows too, since `st_boundary()` returns *every*
ring - the hole's own boundary, not just the outer one - and
`st_length()` sums across all of them. A shape with a hole cut out of
it genuinely is less compact than the solid version, so both terms
moving the same direction is the right behaviour here - but it's not a
universal convention. Some Polsby-Popper implementations in the
redistricting literature measure only the *outer* ring's length,
treating interior holes (a small enclave, a data-digitisation artifact)
as compactness-irrelevant. This package prizes consistency of definition and leaves this data cleaning exercise to
the user.




# Deterministic vs. random estimation

- `convexity_index()`, `span_index()`, `radial_concentration_index()`,
`directional_balance_index()`, and `depth_index()` all have a
`deterministic` argument.`moment_of_inertia_index()`,
- `moment_isotropy_index()`, and the six classic metrics don't have this argument.
They are always computed directly off the CDT mesh / convex hull / bounding
circle.

`deterministic = TRUE` (the default for all five) computes the index
over a **fixed grid** on the CDT mesh - `O(n^2)` in piece count for
convexity/span, `O(n)` for radial_concentration/directional_balance/depth
but with each triangle's own subdivision depth (up to 256 points) scaled to
its area, so only the mesh's largest triangles pay the full cost. It's
called `deterministic`, not
`exact`, deliberately: for shapes with small concavities relative to
triangle size, or (for radial_concentration/directional_balance/depth) an
integral with no closed form at all, it's still only an approximation
of the true value, just a non-random one (see
`vignette("c-understanding-convexity-index")`'s convexity section for
why). 

`deterministic = FALSE` instead computes a **Monte Carlo
estimate**: all five functions draw `n_lines` independent random
samples and average some property of them:
- `convexity_index()` the fraction of a connecting line's length that falls outside the polygon,
- `span_index()` the line's length itself
- `radial_concentration_index()` the distance from each sampled
point to the shape's own geometric median
- `directional_balance_index()` the bearing of each sampled point from the shape's own centroid,
- `depth_index()` the distance from each sampled point to the shape's own
boundary.

Only `convexity_index()` ever actually builds line geometry. Sharing the argument name means one `n_lines` the sample count at once for all the five functions.

```{r}
#| code-fold: true

library(purrr)
library(dplyr)
library(knitr)

# Non-deterministic parameters helper
rand_opts <- list(deterministic = FALSE, n_lines = 3000, seed = 1)

# Helper function to compute metrics for a single shape 
compute_shape_metrics <- function(nm, g) {
  ci_det   <- convexity_index(g, deterministic = TRUE)$index
  ci_rand  <- do.call(convexity_index, c(list(g), rand_opts))$index
  
  span_det  <- span_index(g, deterministic = TRUE)$index
  span_rand <- do.call(span_index, c(list(g), rand_opts))$index
  
  rci_det  <- radial_concentration_index(g, deterministic = TRUE)$index
  rci_rand <- do.call(radial_concentration_index, c(list(g), rand_opts))$index
  
  db_det   <- directional_balance_index(g, deterministic = TRUE)$index
  db_rand  <- do.call(directional_balance_index, c(list(g), rand_opts))$index
  
  depth_det  <- depth_index(g, deterministic = TRUE)$index
  depth_rand <- do.call(depth_index, c(list(g), rand_opts))$index

  tibble::tibble(
    shape               = nm,
    ci_deterministic    = ci_det,
    ci_random_line      = ci_rand,
    span_deterministic  = span_det,
    span_random_pair    = span_rand,
    rci_deterministic   = rci_det,
    rci_random_point    = rci_rand,
    db_deterministic    = db_det,
    db_random_point     = db_rand,
    depth_deterministic = depth_det,
    depth_random_point  = depth_rand,
    moi                 = moment_of_inertia_index(g)$index,
    moment_isotropy     = moment_isotropy_index(g)$index,
    hull_ratio          = hull_ratio_index(g)$index
  )
}

# 1. Loop over shape names to avoid type confusion with `basic_shapes[[nm]]`
det_compare <- map_dfr(names(basic_shapes), function(nm) {
  compute_shape_metrics(nm, basic_shapes[[nm]])
})

# 2. Extract matrix cleanly
det_mat <- as.matrix(det_compare[, -1])
rownames(det_mat) <- det_compare$shape

# 3. Create thumbnail headers safely
headers <- map_chr(names(basic_shapes), function(nm) {
  paste0(shape_thumb(basic_shapes[[nm]]), nm)
})

# 4. Render Table
kable(
  t(det_mat),
  format = "html", 
  digits = 2, 
  row.names = TRUE,
  col.names = headers, 
  escape = FALSE
)
```

The Monte Carlo estimate is noisy, but close to the deterministic one, and that
noise shrinks as `n_lines` grows.

`convexity_index()` always works from the polygon's own CDT triangle
mesh - `n_quad` controls the quadrature refinement on that mesh (see
above), but there's no alternative tessellation to choose. Two standalone
mesh utilities, `convex_decompose()` and `subdivide_mesh()`, are also
available for downstream geometry work (neither is wired into any of the
thirteen indices) - see `vignette("b-understanding-triangulation")` for
both, alongside the CDT mesh they both start from.

# Applying it to an `sf` object, row by row

Real polygons usually arrive as rows of an `sf` object.
`shape_indices_sf()` with `byrow = TRUE` (the default) runs
`shape_indices()` on every row independently and appends four
columns. Two North Carolina counties from the `sf` package - Wake
(a simple, nearly-convex Piedmont county) and Dare (a mainland piece
plus a separated barrier-island strip) - show the range:

```{r nc-byrow}
nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE) %>%
  st_transform(32119)   # NC state plane (meters) - nc.shp ships in NAD27 lon/lat

pair     <- nc[nc$NAME %in% c("Wake", "Dare"), ]
pair_res <- shape_indices_sf(pair)

pair_res %>%
  st_drop_geometry() %>%
  select(NAME, convexity_index, moment_of_inertia_index, span_index, hull_ratio_index) %>%
  knitr::kable(format = "html", digits = 3, row.names = FALSE)
```

```{r nc-byrow-plot, fig.width=7, fig.height=4}
ggplot(pair_res) +
  geom_sf(aes(fill = convexity_index), color = "grey20") +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(fill = "CI")
```

`shape_indices_sf()` also supports `byrow = FALSE`, treating a whole
collection of rows as sub-polygons of one weighted overall shape rather
than scoring each row independently - and weighting them changes the
CDT mesh itself, not just the resulting index values, down to a design
decision that treats a row weighted `0` or `NA` as an excluded hole
rather than a zero-weighted piece. See
`vignette("b-understanding-triangulation")` for both mechanics, and
`vignette("k-nc-counties-comparison")`'s "Weighting a collection of
polygons" for how the weighting choice moves the index values
themselves.

# Parallelisation and efficiency

This package has one parallelism knob, at the row level:
`shape_indices_sf()`'s `parallel_rows` argument. There is no
parallelism *within* a single index call - a single very complex
polygon can't be split across cores, only a collection of rows can.

With `byrow = TRUE`, `parallel_rows = TRUE` distributes whole rows
across the active `future::plan()` via `furrr` - this dispatch is
unconditional (no fast-path short-circuit), so it delivers real
multi-core speedups. Row-level parallelism is only worth turning on
once each row's own cost is bounded - pass `deterministic_max_tri` (see
`shape_indices()`) so no single county can blow up into an expensive
deterministic-mode computation (`O(n^2)` for convexity/span, `O(n)`
with a large constant for radial_concentration) inside a worker.
Raising `deterministic_max_tri` and
`n_lines` below (relative to the row-by-row example earlier) gives every
county enough work to make the per-worker overhead worth paying:

```{r parallel-rows}
library(furrr)

future::plan(future::sequential)
t_seq_rows <- system.time(
  res_seq <- shape_indices_sf(nc, deterministic_max_tri = 60, n_lines = 6000, seed = 1,
                               parallel_rows = FALSE)
)

future::plan(future::multisession, workers = 4)
t_par_rows <- system.time(
  res_par_rows <- shape_indices_sf(nc, deterministic_max_tri = 60, n_lines = 6000, seed = 1,
                                    parallel_rows = TRUE)
)
future::plan(future::sequential)

timing <- data.frame(
  mode    = c("parallel_rows = FALSE", "parallel_rows = TRUE (4 workers)"),
  elapsed = c(t_seq_rows[["elapsed"]], t_par_rows[["elapsed"]])
)
timing$speedup <- timing$elapsed[1] / timing$elapsed
knitr::kable(format = "html", timing, digits = 2, row.names = FALSE)
```


Both modes agree on every county's indices (not shown - parallelising
changes *how* the 100 rows get computed, not the values themselves);
the gain is in `elapsed`. Overhead from spinning up workers is fixed
per call, so it matters less as the batch gets bigger or each row
gets more expensive - not worth turning on for a handful of small,
cheap polygons, but pays off exactly on the kind of workload this
package targets: many real-world rows, each potentially complex enough
to need `deterministic_max_tri`'s random-line fallback. For a real workload,
size `workers` to `future::availableCores()` (or fewer, to leave
headroom) and prefer `future::multicore` over `future::multisession`
on Mac/Linux, which forks rather than starting fresh R sessions per
worker.

# `simplify_tolerance`: reducing boundary detail, not just mesh size

`prepare_polygon()`, `shape_indices()`, and `shape_indices_sf()` all
accept a `simplify_tolerance` argument: an `st_simplify()` pass, applied
once the polygon is already cleaned and (for `byrow = FALSE`) already
unioned, before triangulating. Its units are whatever the geometry's own
CRS uses at that point - metres for geographic (lon/lat) input, since
these functions auto-project that to a metric CRS themselves, but
whatever unit an *already*-projected input's CRS uses otherwise (many US
State Plane CRSs are US survey feet, not metres) - a warning fires if
that unit isn't metres, since the same numeric tolerance means a very
different amount of simplification depending on which.

It's worth reaching for on genuinely
detailed real-world boundaries (block/parcel data with hundreds of
thousands of vertices), for a reason that isn't obvious from mesh size
alone: `convexity_index()`'s `deterministic = FALSE` mode tests
candidate lines against *every* boundary edge of the polygon, and that
edge count doesn't shrink just because the triangulated mesh does -
simplifying the boundary itself is what actually bounds it.

**With `byrow = FALSE`, simplify the union, never the rows
beforehand.** `shape_indices_sf(byrow = FALSE)` applies
`simplify_tolerance` to `st_union(x)` itself, after every row's own
boundary has already merged with its neighbours - internal edges
shared between adjacent rows are already gone by that point, so there's
nothing there for simplification to misalign. Simplifying each row
independently *first* - a natural-looking alternative, and the wrong
one - simplifies the same shared edge differently on either side of
it, turning what used to be a clean tessellation into thousands of
sliver gaps between rows that no longer quite touch. As a concrete
example, on one real 50,424-row Census-block dataset: unioning the raw
rows gives one clean, ~50,000-vertex polygon, while independently
simplifying every row first and *then* unioning gives 1,851 disjoint
sliver fragments instead - *more* total boundary than not simplifying
at all, which at that scale is the difference between a computation
that fits in memory and one that doesn't. Passing `simplify_tolerance`
here instead keeps the union at one part and cuts its boundary edge
count by an order of magnitude. Weight accuracy is unaffected by
design: the row-level weight overlay always uses the *unsimplified*
original rows - only the mesh comes from the simplified union - so a
tolerance modest relative to the shape's own scale still captures
upward of 99% of the true row-level weight total.
