---
title: "5. Understanding Radial Concentration Index"
author: "Nikhil Kaza"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{5. Understanding Radial Concentration Index}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
  width: auto !important;
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.align = "center", cache = TRUE)
```

```{r load, cache=FALSE, echo=TRUE}
#| code-fold: true
library(shapeindices)
library(sf)
library(ggplot2)

theme_set(theme_minimal(base_size = 11))
theme_gallery <- theme_void(base_size = 10) +
  theme(strip.text = element_text(size = 9, face = "bold"))
```

```{r shapes, echo=TRUE}
#| code-fold: true
square <- st_polygon(list(rbind(c(0,0), c(10,0), c(10,10), c(0,10), c(0,0))))
make_rect <- function(w, h) st_polygon(list(rbind(c(0,0), c(w,0), c(w,h), c(0,h), c(0,0))))
aspect_seq <- c(1, 2, 4, 10, 20)
rectangles <- lapply(aspect_seq, function(a) make_rect(sqrt(100 * a), sqrt(100 / a)))
names(rectangles) <- sprintf("aspect %gx", aspect_seq)

make_star <- function(n_points, r_outer = 1, r_inner = 0.5, center = c(0, 0)) {
  n <- n_points * 2
  angles <- seq(pi/2, pi/2 + 2*pi, length.out = n + 1)[1:n]
  radii  <- rep(c(r_outer, r_inner), n_points)
  x <- center[1] + radii * cos(angles); y <- center[2] + radii * sin(angles)
  st_polygon(list(rbind(cbind(x, y), c(x[1], y[1]))))
}
ratio_seq <- c(0.9, 0.7, 0.5, 0.3, 0.15)
stars_r <- lapply(ratio_seq, function(r) make_star(6, 5, 5 * r))
names(stars_r) <- sprintf("notch ratio %.2f", ratio_seq)

disk <- st_buffer(st_sfc(st_point(c(0, 0))), dist = 5.64, nQuadSegs = 60)[[1]]

# a ring of n_arms radial wedges spanning [r_in, r_out], with a FIXED total
# angular width split evenly among them, so n_arms only changes how finely
# that fixed width is cut up - same construction used in
# vignette("c-understanding-moment-of-inertia-index") to demonstrate the
# same underlying effect there
make_spokes <- function(r_in, r_out, n_arms, total_angle_frac = 0.5) {
  angles <- seq(0, 2 * pi, length.out = n_arms + 1)[1:n_arms]
  half_w <- total_angle_frac * pi / n_arms
  polys <- lapply(angles, function(a0) {
    th <- seq(a0 - half_w, a0 + half_w, length.out = max(6, 40 %/% n_arms))
    outer_pts <- cbind(r_out * cos(th), r_out * sin(th))
    inner_pts <- cbind(r_in * cos(rev(th)), r_in * sin(rev(th)))
    st_polygon(list(rbind(outer_pts, inner_pts, outer_pts[1, ])))
  })
  Reduce(function(p, q) st_union(st_sfc(p), st_sfc(q))[[1]], polys)
}
arm_counts <- c(2, 4, 8, 16)
spoke_shapes <- lapply(arm_counts, function(n) make_spokes(3, 5, n))
names(spoke_shapes) <- sprintf("%d arms", arm_counts)
```

# Introduction

`radial_concentration_index()` measures how tightly a polygon's own mass is concentrated around its own best possible centre, relative to how concentrated an equal-area circle's mass is around *its* centre. For a mass distribution with density $\rho$ and total mass $W$, and a candidate centre $c$,

$$D_1(\rho, c) = \frac{1}{W}\int_P \rho(s)\,\lvert s - c\rvert\,ds, \qquad D_1(\rho) = \min_{c} D_1(\rho, c)$$

As in `span_index()`, $\rho$ is a density over the polygon and $s \in P$ is a whole point, not a scalar coordinate. But here the density appears *once*, not twice as in span's $\rho(s)\,\rho(t)$ double integral, because only one end of the distance is a random point: $s$ is drawn from $\rho/W$, while $c$ is a single fixed candidate centre, not a second draw. That doesn't mean $\rho$ acts only once - the minimiser $c$ is itself the $\rho$-weighted geometric median, so reweighting the mass also moves the centre - it just enters through the $\min_{c}$ rather than as a second integration variable.

`radial_concentration_index()` is $D_{1,\text{ref}}/D_1(\rho) \in (0, 1]$, where $D_{1,\text{ref}}$ is the same quantity for a circle (unweighted) or a concentric annulus (weighted) of the same area. It's a sibling to `moment_of_inertia_index()` (squared distance to the centroid) and `span_index()` (plain distance, but between two random points, no reference point at all) - this one is plain distance *to* a reference point, the remaining combination of the two design choices.

No single prior paper matches this exact construction - median-anchored, plain-distance dispersal - the way `moment_of_inertia_index()` matches both Kaza's IMI and Angel et al.'s Cohesion (`vignette("c-understanding-moment-of-inertia-index")`). It belongs generically to the "dispersion" family of circle-compactness properties surveyed in Angel, Parent & Civco (2010)[^angel-e] - shapes whose mass sits close to a well-chosen centre read as compact, wherever exactly that centre is anchored - without being a direct implementation of any one property in that survey.

[^angel-e]: Angel, S., Parent, J., & Civco, D. L. (2010). Ten compactness properties of circles: measuring shape in geography. *Canadian Geographer*, 54(4), 441-461.

```{r glance-helper}
# a small base64-embedded PNG of a shape, for use as a table cell in place
# of a text label - knitr::image_uri() does the embedding, so no new
# package dependency (it's already pulled in transitively via knitr, which
# every vignette in this package already Suggests)
shape_thumb <- function(geom, size_px = 40, col = "grey75") {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(st_geometry(st_sfc(geom)), col = col, border = "grey30")
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}

# same idea, but shaded by a triangle mesh's own per-triangle weight instead
# of a flat fill - for tables comparing weight DISTRIBUTIONS on one fixed
# shape rather than comparing different shapes
weight_thumb <- function(tri, w, size_px = 40) {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  pal  <- grDevices::colorRampPalette(c("grey95", "firebrick"))(100)
  cols <- pal[as.integer(cut(w, 100, labels = FALSE))]
  plot(st_geometry(tri), col = cols, border = NA)
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}
```

# Deriving the index

## Why the geometric median, not the centroid

$D_1(\rho, c)$ is convex in $c$, so it has a unique minimiser - but it's the **geometric median** (Fermat-Weber point), not the centroid. The centroid minimises *squared* distance ($\mathbb E\lvert s - c\rvert^2$, which is why `moment_of_inertia_index()` uses it); plain distance is minimised by a different point entirely, with no closed form of its own. The two coincide only when the shape is symmetric enough (a disk, or anything centrally symmetric); on a lopsided shape they separate, with the median pulled toward the bulk of the mass and the centroid sitting at the plain average of position.

The figure below shows both on a "tadpole" - a round head with a thin tail, deliberately lopsided - together with the spokes whose lengths $D_1$ averages, drawn from the median (the point that makes that average as small as possible).

```{r median-vs-centroid, fig.width=7, fig.height=4}
tadpole <- st_union(
  st_buffer(st_sfc(st_point(c(0, 0))), 4, nQuadSegs = 40),
  st_sfc(st_polygon(list(rbind(
    c(0, -0.6), c(16, -0.6), c(16, 0.6), c(0, 0.6), c(0, -0.6))))))

rc  <- radial_concentration_index(tadpole)
med <- st_coordinates(rc$center)[1, 1:2]
cen <- st_coordinates(st_centroid(tadpole))[1, 1:2]

set.seed(3)
samp <- st_coordinates(st_sample(tadpole, size = 45, type = "random", exact = TRUE))[, 1:2]
spokes <- data.frame(x = med[1], y = med[2], xend = samp[, 1], yend = samp[, 2])
pts <- data.frame(
  x = c(med[1], cen[1]), y = c(med[2], cen[2]),
  label = c("geometric median (minimises mean distance)", "area centroid"))

ggplot() +
  geom_sf(data = tadpole, fill = "grey90", color = "grey40") +
  geom_segment(data = spokes, aes(x, y, xend = xend, yend = yend),
               color = "steelblue", linewidth = 0.3, alpha = 0.7) +
  geom_point(data = pts, aes(x, y, shape = label, color = label), size = 4, stroke = 1.5) +
  scale_shape_manual(values = c(8, 3), name = NULL) +
  scale_color_manual(values = c("firebrick", "grey20"), name = NULL) +
  coord_sf() +
  theme_gallery +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

The tail drags the area centroid off toward it, while the geometric median stays back in the head where most of the mass actually is - the robustness that makes the median, not the centroid, the right centre for a mean-distance measure. The index is the mean spoke length from that median: the smallest such average any single centre can achieve. On a symmetric shape the two markers coincide and the distinction vanishes.

## The disk and annulus are provably optimal

For a *fixed* centre $c$, the disk of area $A$ centred at $c$ minimises $\int_\Omega|s-c|\,ds$ among every measurable set $\Omega$ of that area - a direct bathtub-principle argument ($|s-c|$ is radially increasing about $c$, so the minimising set of fixed measure is a ball around it). Evaluating this at $\Omega$'s own geometric median shows the disk beats *every* shape of that area, holes and multiple parts included - the proof never assumes connectivity or convexity. The weighted case gives concentric rings, sorted densest-to-centre, by the same argument applied level-set by level-set.

## Closed forms

Distance-to-a-fixed-point depends only on radius, not angle, so - unlike `span_index()` - neither reference needs elliptic integrals:

$$D_{1,\text{circle}}(A) = \frac{2}{3}\sqrt{\frac{A}{\pi}}, \qquad \bar r(a, b) = \frac{2(a^2+ab+b^2)}{3(a+b)} \text{ for a ring of inner/outer radius } a, b$$

## Finding the centre

The geometric median has no closed form, so it's found via **Weiszfeld's algorithm**: starting from the centroid, repeatedly re-weight each point by the inverse of its current distance to the estimate and re-average, until the *objective value* (not the position) stops moving. Convergence on position isn't guaranteed for a symmetric multi-part shape - two identical, separated blobs have a whole segment of equally-good centres, not one point - so the stopping rule has to watch the value, not the point.

```{r weiszfeld-demo, echo=TRUE}
sq1 <- st_polygon(list(rbind(c(0,0), c(2,0), c(2,2), c(0,2), c(0,0))))
sq2 <- st_polygon(list(rbind(c(10,0), c(12,0), c(12,2), c(10,2), c(10,0))))
dumbbell <- st_union(st_sfc(sq1, sq2))
res <- radial_concentration_index(dumbbell)
st_coordinates(res$center)   # lands exactly on the connecting midpoint, (6, 1)
```

Each triangle also needs finer internal resolution than a fixed 3-point quadrature rule gives - collapsing a large triangle to a handful of fixed points systematically misestimates $\int_T|s-c|\,ds$, by single digits of percent at realistic mesh sizes, not a rounding-level nicety. `radial_concentration_index()`'s default (`deterministic = TRUE`) instead recursively subdivides each triangle - the same medial-subdivision idea `span_index()` uses for its own self-term, though built by its own vectorised routine that subdivides every triangle in the mesh in one pass rather than recursing triangle-by-triangle - into up to 256 sub-triangles and uses their centroids as the point cloud. The depth is area-adaptive, not fixed: the bias above is worst for *large* triangles specifically (it scales with a triangle's own spatial extent), so only the mesh's own largest triangle pays the full 256 points - smaller triangles need proportionally less subdivision to reach that same final sub-triangle resolution, down to a single centroid for the smallest ones. A real-world CDT mesh - many small triangles packed along complex boundary detail alongside a handful of large ones in the open interior - can have area ratios of many orders of magnitude between its largest and smallest triangle, so this typically cuts the total point count by 60x or more relative to giving every triangle the same fixed depth, while changing the index by under 0.05% - invisible at the precision indices are typically read to. `deterministic = FALSE` remains the escape valve for meshes too large even after that: it samples `n_lines` points directly from the weighted density and runs Weiszfeld on that sample instead, the same Monte Carlo idea `convexity_index()`/`span_index()` already use for the same reason.

```{r rci-mc-demo, echo=TRUE}
deep_star <- stars_r[["notch ratio 0.15"]]
radial_concentration_index(deep_star)$index
radial_concentration_index(deep_star, deterministic = FALSE, n_lines = 3000, seed = 1)$index
```

# Illustrations

## Basic shapes

```{r basic-table}
rows <- c(list(square = square, disk = disk), rectangles, stars_r)
tbl <- do.call(rbind, lapply(names(rows), function(nm) {
  r <- radial_concentration_index(rows[[nm]])
  data.frame(shape = shape_thumb(rows[[nm]]), name = nm,
             radial_concentration_index = r$index, D1 = r$D1, D1_ref = r$D1_ref)
}))
knitr::kable(format = "html", tbl, digits = 3, row.names = FALSE, escape = FALSE)
```

The disk scores just above 1 rather than exactly 1 - the expected residual from finite subdivision depth, not a modelling error. Elongation and deeper notches both pull the index down, same direction as `moment_of_inertia_index()` and `span_index()`, but not identically - it's measuring a different thing (mean distance *to a point*, not squared distance to the centroid or pairwise distance with no reference point at all).

## Holes and multi-part shapes

```{r hole-multipart, fig.width=8, fig.height=4}
outer <- rbind(c(-5,-5), c(5,-5), c(5,5), c(-5,5), c(-5,-5))
hole  <- rbind(c(-2,-2), c(-2,2), c(2,2), c(2,-2), c(-2,-2))
square_hole <- st_polygon(list(outer, hole))
res_hole <- radial_concentration_index(square_hole)

res_dumbbell <- radial_concentration_index(dumbbell)

combined <- rbind(
  st_sf(label = sprintf("square with hole (index = %.3f)", res_hole$index), geometry = st_sfc(square_hole)),
  st_sf(label = sprintf("two separated squares (index = %.3f)", res_dumbbell$index), geometry = st_sfc(dumbbell))
)
centres <- rbind(
  st_sf(label = combined$label[1], geometry = res_hole$center),
  st_sf(label = combined$label[2], geometry = res_dumbbell$center)
)

ggplot() +
  geom_sf(data = combined, fill = "grey85", color = "grey20") +
  geom_sf(data = centres, color = "firebrick", size = 2) +
  facet_wrap(~ label) +
  theme_gallery
```

Both centres (red points) are found correctly even though neither sits on solid material - one lands in the middle of the hole, the other on the empty gap between the two squares. This is expected: the geometric median, like the ordinary centroid, doesn't have to be a point of the shape itself, and the bathtub-principle proof never assumed it would be.

## Things to watch out for

### Same radial spread, same index

`radial_concentration_index()` shares `moment_of_inertia_index()`'s exact insensitivity to angular rearrangement, for the same underlying reason (see `vignette("c-understanding-moment-of-inertia-index")` for the full derivation): $D_1 = \int_P \rho(s)\,|s-c|\,ds$ depends on each point only through its distance from the reference centre $c$, never its bearing. For a shape with two-fold or higher rotational symmetry, $c$ - here the geometric median, not the centroid - sits at the shape's own centre of symmetry regardless of how finely the mass is carved up angularly, so the index can't tell a couple of broad wedges from many thin ones covering the same total angle.

```{r blind-spot, echo=TRUE}
spoke_results <- lapply(spoke_shapes, function(g) radial_concentration_index(st_sfc(g)))
tbl_spokes <- do.call(rbind, Map(function(nm, res) {
  data.frame(shape = shape_thumb(spoke_shapes[[nm]]), name = nm,
             radial_concentration = res$index,
             convexity = suppressWarnings(convexity_index(st_sfc(spoke_shapes[[nm]]), deterministic = FALSE,
                                                            n_lines = 5000, seed = 1)$index))
}, names(spoke_shapes), spoke_results))
knitr::kable(format = "html", tbl_spokes, digits = 4, row.names = FALSE, escape = FALSE)

# how far each row's own median landed from the origin - checking that the
# rotational symmetry really is pinning it there, not just approximately
center_drift <- vapply(spoke_results, function(res) {
  as.numeric(sqrt(sum(st_coordinates(res$center)[1, 1:2]^2)))
}, numeric(1))
```

`radial_concentration` doesn't move as the same fixed angular width is cut from 2 arms into 16 - only `convexity_index()` responds to the widening gaps. Worth checking independently here, since the reference centre is a median rather than a centroid: the largest distance from the origin to any row's own `center` above is `r sprintf("%.2e", max(center_drift))`, confirming the median stays pinned by the same rotational symmetry the centroid relies on in `moment_of_inertia_index()`'s own version of this effect.

### Weight moves the reference, not just the value

```{r weighted-example}
star <- make_star(8, 5, 2)
prep <- prepare_polygon(star)
cen <- st_coordinates(st_centroid(st_union(prep$poly)))[1, 1:2]
tc  <- st_coordinates(st_centroid(st_geometry(prep$tri)))
d   <- sqrt((tc[,1]-cen[1])^2 + (tc[,2]-cen[2])^2)

w_center <- exp(-d / mean(d))
w_edge   <- exp(d / mean(d))
data.frame(
  weighting = c(weight_thumb(prep$tri, rep(1, nrow(prep$tri))),
                weight_thumb(prep$tri, w_center), weight_thumb(prep$tri, w_edge)),
  name = c("uniform", "concentrated at centre", "concentrated at edge"),
  index = c(radial_concentration_index(star, prep = prep)$index,
            radial_concentration_index(star, prep = prep, weight = w_center)$index,
            radial_concentration_index(star, prep = prep, weight = w_edge)$index)
) |> knitr::kable(format = "html", digits = 3, row.names = FALSE, escape = FALSE)
```

Concentrating weight at the edge scores lowest, as expected. But concentrating it at the centre scores *lower than the uniform baseline*, not higher - surprising at first sight, until you remember that the reference itself adapts to the given weight's own histogram rather than staying fixed. Both `D1` (actual) and `D1_ref` shrink when weight moves toward the centre, but here `D1_ref` shrinks proportionally faster: pushing weight toward the middle via a smooth radial function creates a skewed histogram whose *own* ideal sorted-ring arrangement is a much tighter target than a uniform disk's, and the star's actual notches keep the real arrangement from fully reaching that tighter target. The one comparison guaranteed to hold either way - centre-weighted beats edge-weighted - does; the comparison to the unweighted baseline doesn't have a similarly simple guarantee.

# Where this fits

`radial_concentration_index()` is the plain-distance-to-a-reference-point member of a family of three. Its siblings measure the same "how spread out is the mass" question through different lenses: `moment_of_inertia_index()` uses *squared* distance to the centroid (`vignette("c-understanding-moment-of-inertia-index")`), and `span_index()` uses plain distance between two random points, with no fixed reference point at all (`vignette("d-understanding-span-index")`). Unlike either sibling, its reference centre is a geometric median rather than a centroid - the point that makes plain (not squared) distance minimal - which is also why it, alone among the three, needs an iterative algorithm (Weiszfeld's) just to locate its own centre.

"Things to watch out for" above already showed one divergence from `convexity_index()`: widening the *angular* gaps between spokes at a fixed radius moves `convexity_index()` but leaves `radial_concentration_index()` untouched. Pulling two solid pieces apart in *space* - rather than carving a fixed footprint more finely - shows the opposite pattern, and a genuinely surprising one: `radial_concentration_index()` turns out to be the more sensitive of the two, not the less.

```{r rci-vs-convexity, fig.width=7, fig.height=4.5}
make_dumbbell_gap <- function(gap) {
  sq1 <- st_polygon(list(rbind(c(0,0), c(2,0), c(2,2), c(0,2), c(0,0))))
  sq2 <- st_polygon(list(rbind(c(2+gap,0), c(4+gap,0), c(4+gap,2), c(2+gap,2), c(2+gap,0))))
  st_union(st_sfc(sq1, sq2))
}
gaps <- c(0.5, 2, 5, 10, 20, 40)
gap_cmp <- do.call(rbind, lapply(gaps, function(gp) {
  g <- make_dumbbell_gap(gp)
  data.frame(gap = gp,
             radial_concentration = radial_concentration_index(g)$index,
             convexity = suppressWarnings(convexity_index(g, deterministic = FALSE, n_lines = 5000, seed = 1)$index))
}))
gap_long <- data.frame(
  gap = rep(gap_cmp$gap, 2),
  index_name = rep(c("radial_concentration_index", "convexity_index"), each = nrow(gap_cmp)),
  value = c(gap_cmp$radial_concentration, gap_cmp$convexity))

ggplot(gap_long, aes(gap, value, color = index_name)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_color_manual(values = c("steelblue", "firebrick"), name = NULL) +
  labs(x = "gap between the two squares", y = "index value") +
  ylim(0, 1) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

Both indices fall as the two squares separate, but only one of them keeps falling. `convexity_index()` levels off near `r sprintf("%.2f", gap_cmp$convexity[nrow(gap_cmp)])` no matter how far apart the pieces get - it's a bounded probability (the fraction of random point-pairs whose connecting line stays inside), and once the two lobes are far enough apart that essentially every cross-lobe pair is already cut, further separation can't push the fraction down any more. `radial_concentration_index()` has no such floor: it's an absolute mean-distance ratio, and mean distance from the (unboundedly distant) median to either lobe grows without limit as the gap widens, so the index keeps falling toward 0. The lesson generalises beyond this one shape: `convexity_index()` asks a yes/no question per pair of points and averages the answers, which saturates once the answer stops changing; the "distance from a centre" family (`radial_concentration_index()`, `moment_of_inertia_index()`, `span_index()`) has no equivalent ceiling on how badly a shape can score.

# Key takeaways

- `radial_concentration_index()` compares $D_1$ (mean plain distance from every point to the shape's own geometric median) to the same quantity for an equal-area disk, or - weighted - a concentric annulus, both provably optimal via the bathtub principle. 1 exactly when the shape matches its reference.
- It anchors on the geometric median, not the centroid: the point that minimises plain distance, found only iteratively (Weiszfeld's algorithm), and one that can diverge sharply from the centroid on a lopsided shape (see the tadpole figure above).
- It depends only on distance from that centre, never bearing: shapes with two-fold or higher rotational symmetry and the same radius-vs-mass profile score *identically*, however finely that mass is carved into separate wedges - the same effect `moment_of_inertia_index()` has, and for the same reason (see "Things to watch out for" above). `convexity_index()` is the one that actually sees the gaps between the wedges.
- Weighting doesn't just change the value, it can change *which shape wins*: concentrating weight toward the centre can score *lower* than uniform weighting, because the reference itself tightens faster than the actual value does (see "Things to watch out for" above) - the one guarantee that survives is that centre-weighting always beats edge-weighting on the same shape.
- Holes and multi-part gaps are handled correctly by construction: the median can (and does) land on empty territory, exactly like the centroid, since the bathtub-principle proof never assumed the reference point had to sit on solid material.
- `deterministic = TRUE` (the default) uses area-adaptive triangle subdivision rather than a fixed quadrature rule, since a small, fixed point count systematically misestimates $\int_T|s-c|\,ds$ on real meshes; `deterministic = FALSE` remains available as a Monte Carlo escape valve on meshes too large even for that.
