---
title: "5. Understanding Span Index"
author: "Nikhil Kaza"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{5. Understanding Span Index}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
  width: auto !important;
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.align = "center", cache = TRUE)
```

```{r load, cache=FALSE, echo=TRUE}
#| code-fold: true
library(shapeindices)
library(sf)
library(ggplot2)
library(dplyr)

theme_set(theme_minimal(base_size = 11))
theme_gallery <- theme_void(base_size = 10) +
  theme(strip.text = element_text(size = 9, face = "bold"),
        legend.position = "bottom")
```

```{r shapes, echo=TRUE}
#| code-fold: true
square <- st_polygon(list(rbind(c(0,0), c(10,0), c(10,10), c(0,10), c(0,0))))

make_rect <- function(w, h) st_polygon(list(rbind(c(0,0), c(w,0), c(w,h), c(0,h), c(0,0))))
aspect_seq  <- c(1, 2, 4, 10, 20)
rectangles  <- lapply(aspect_seq, function(a) make_rect(sqrt(100 * a), sqrt(100 / a)))
names(rectangles) <- sprintf("aspect %gx", aspect_seq)

make_star <- function(n_points, r_outer = 1, r_inner = 0.5, center = c(0, 0)) {
  n <- n_points * 2
  angles <- seq(pi/2, pi/2 + 2*pi, length.out = n + 1)[1:n]
  radii  <- rep(c(r_outer, r_inner), n_points)
  x <- center[1] + radii * cos(angles)
  y <- center[2] + radii * sin(angles)
  coords <- rbind(cbind(x, y), c(x[1], y[1]))
  st_polygon(list(coords))
}
ratio_seq <- c(0.9, 0.7, 0.5, 0.3, 0.15)
stars_r   <- lapply(ratio_seq, function(r) make_star(6, 5, 5 * r))
names(stars_r) <- sprintf("notch ratio %.2f", ratio_seq)

disk <- st_buffer(st_sfc(st_point(c(0, 0))), dist = 5.64, nQuadSegs = 60)[[1]]  # area matches square

# a ring of n_arms radial wedges spanning [r_in, r_out], with a FIXED total
# angular width split evenly among them, so n_arms only changes how finely
# that fixed width is cut up - same construction used in
# vignette("d-understanding-moment-of-inertia-index") and
# vignette("f-understanding-radial-concentration-index") to demonstrate
# those two indices' shared exact invariance to this kind of rearrangement
make_spokes <- function(r_in, r_out, n_arms, total_angle_frac = 0.5) {
  angles <- seq(0, 2 * pi, length.out = n_arms + 1)[1:n_arms]
  half_w <- total_angle_frac * pi / n_arms
  polys <- lapply(angles, function(a0) {
    th <- seq(a0 - half_w, a0 + half_w, length.out = max(6, 40 %/% n_arms))
    outer_pts <- cbind(r_out * cos(th), r_out * sin(th))
    inner_pts <- cbind(r_in * cos(rev(th)), r_in * sin(rev(th)))
    st_polygon(list(rbind(outer_pts, inner_pts, outer_pts[1, ])))
  })
  Reduce(function(p, q) st_union(st_sfc(p), st_sfc(q))[[1]], polys)
}
arm_counts <- c(2, 4, 8, 16)
spoke_shapes <- lapply(arm_counts, function(n) make_spokes(3, 5, n))
names(spoke_shapes) <- sprintf("%d arms", arm_counts)
```

# Introduction

`span_index()` measures how far apart a polygon's own interior points typically are, relative to a reference shape. Two independent points $s, t \in P$ drawn uniformly from a polygon $P$ have some mean distance apart, $D(P) = \mathbb{E}|s - t|$; `span_index()` compares this to $D_{\text{ref}}$, the same quantity for the *most compact possible* arrangement of that area/mass - a circle when unweighted, a concentric annulus (densest at the centre) when weighted. The index is $D_{\text{ref}}/D(P) \in (0, 1]$, equal to 1 exactly when $P$ already matches its reference.

This sits alongside `convexity_index()` (which measures whether interior lines stay inside the boundary) and `moment_of_inertia_index()` (which measures dispersal via *squared* distance from the shape's own centroid). Span index measures dispersal too, but via plain (non-squared) pairwise distance - a genuinely different quantity, not just a rescaling of moment of inertia, for reasons this vignette derives.

Span index's plain-distance construction is close to, but not quite, a named index in the shape-compactness literature (reviewed generally in Frolov 1975[^frolov] and MacEachren 1985[^maceachren]). Angel, Parent & Civco (2010)[^cohesion] define **Cohesion** - one of several classical "compactness properties of circles," alongside Polsby-Popper and Reock (both already in this package, as `polsby_popper_index()` and `reock_index()`) - as the ratio of mean *squared* distance between two random points in an equal-area circle to the same quantity in the actual shape. That's `moment_of_inertia_index()`'s own construction, not span index's: mean squared pairwise distance is exactly twice the mean squared distance to the centroid for any point set, and that factor of 2 cancels identically in the ratio, so Cohesion reduces to Kaza's IMI exactly (see `vignette("d-understanding-moment-of-inertia-index")`'s Introduction). Angel et al. explicitly flag the *plain*-distance version - span index's own construction - as an open question in their own footnote: "It stands to reason that among all shapes of a given area the circle also has the shortest average distance...but a rigorous proof of this proposition has so far eluded us." The rearrangement-inequality argument below (see "The reference shapes really are optimal") settles exactly that question.

[^frolov]: Frolov, Y. S. (1975). Measuring the shape of geographical phenomena: a history of the issue. *Soviet Geography*, 16(10), 676-687.

[^maceachren]: MacEachren, A. M. (1985). Compactness of geographic shape: comparison and evaluation of measures. *Geografiska Annaler: Series B, Human Geography*, 67(1), 53-67.

[^cohesion]: Angel, S., Parent, J., & Civco, D. L. (2010). Ten compactness properties of circles: measuring shape in geography. *Canadian Geographer*, 54(4), 441-461.

```{r glance-helper}
# a small base64-embedded PNG of a shape, for use as a table cell in place
# of a text label - knitr::image_uri() does the embedding, so no new
# package dependency (it's already pulled in transitively via knitr, which
# every vignette in this package already Suggests)
shape_thumb <- function(geom, size_px = 40, col = "grey75") {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(st_geometry(st_sfc(geom)), col = col, border = "grey30")
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}
```

# Deriving span index

## Definition

Here $\rho$ is a density over the polygon $P$: for a point $s \in P$, $\rho(s)$ is the mass density there. Let $s, t \in P$ be two independent points drawn from $\rho/W$, where $W = \int_P \rho$ is the total mass. Define

$$D(\rho) = \frac{1}{W^2}\iint_{P \times P} \rho(s)\,\rho(t)\,\lvert s - t\rvert\; ds\, dt = \mathbb{E}\,\lvert s - t\rvert$$

The density appears *twice* - once for each of the two independent points $s$ and $t$ - because span index is about a pair of random locations (this is what distinguishes it from `radial_concentration_index()`, where only one end of the distance is random and $\rho$ appears once). The letters $s, t$ stand for whole points of $P$, not scalar coordinates.

Span index is $D_{\text{ref}}/D(\rho)$, where $D_{\text{ref}}$ is this same quantity computed for a reference mass arrangement that provably minimises it (below). Unweighted, $\rho$ is just the polygon's indicator function and $W = A$ (area); weighted, $\rho$ is piecewise-constant across the CDT triangles, one `weight` value per triangle, and $W = \sum \text{weight}$.

## What the index sees

Span index is the *average length of the segment between two random interior points* - a chord, strictly, has both endpoints on the boundary itself, so "segment" is the accurate word here even though the two are easy to conflate. Draw a few dozen such segments and colour them by length: a compact shape produces mostly short ones, an elongated one of the same area produces many long ones, so its mean pairwise distance $D$ is larger and its index (reference over $D$) is lower. The dashed circle in each panel is that shape's own equal-area disk reference - the two shapes' interior points would be, on average, the same distance apart only if the panel's own polygon were itself a disk.

```{r what-it-sees, fig.width=11, fig.height=4.5}
set.seed(1)
random_segments <- function(geom, n = 45) {
  pts <- st_coordinates(st_sample(st_sfc(geom), size = 2 * n, type = "random", exact = TRUE))
  pts <- pts[seq_len(2 * n), , drop = FALSE]
  a <- pts[1:n, , drop = FALSE]; b <- pts[(n + 1):(2 * n), , drop = FALSE]
  data.frame(x = a[, 1], y = a[, 2], xend = b[, 1], yend = b[, 2],
             len = sqrt((a[, 1] - b[, 1])^2 + (a[, 2] - b[, 2])^2))
}

# geom_sf() can't do free per-facet scales, so normalise each shape onto a
# common local bounding box (centred, longest side scaled to 1) before
# faceting - same fix used in vignette("c-understanding-convexity-index") -
# otherwise a shared fixed scale across all three panels (the square, the
# 10x-aspect rectangle spanning 30+ units, and the origin-centred star)
# would squash the compact shapes down to slivers and leave them off-centre
local_frame <- function(geom) {
  bb <- st_bbox(st_sfc(geom))
  list(center = c((bb["xmin"] + bb["xmax"]) / 2, (bb["ymin"] + bb["ymax"]) / 2),
       scale  = 1 / max(bb["xmax"] - bb["xmin"], bb["ymax"] - bb["ymin"]))
}
normalize_geom <- function(geom, fr) (geom - fr$center) * fr$scale
normalize_xy    <- function(x, y, fr) list(x = (x - fr$center[1]) * fr$scale, y = (y - fr$center[2]) * fr$scale)

demo <- list(compact = square, elongated = rectangles[["aspect 10x"]], notched = make_star(6, 5, 1.5))
panel_of <- function(nm) sprintf("%s  (span index %.2f)", nm, span_index(demo[[nm]])$index)

segment_df <- do.call(rbind, lapply(names(demo), function(nm) {
  fr <- local_frame(demo[[nm]])
  d  <- random_segments(demo[[nm]])
  a  <- normalize_xy(d$x, d$y, fr); b <- normalize_xy(d$xend, d$yend, fr)
  # segment length keeps its TRUE (pre-normalisation) value for the colour
  # scale, so the elongated rectangle's segments still read as genuinely
  # longer, not rescaled away to fit the same panel as everything else
  data.frame(x = a$x, y = a$y, xend = b$x, yend = b$y, len = d$len, panel = panel_of(nm))
}))
poly_df <- do.call(rbind, lapply(names(demo), function(nm) {
  fr <- local_frame(demo[[nm]])
  data.frame(panel = panel_of(nm), geom = st_as_text(normalize_geom(st_sfc(demo[[nm]]), fr)))
}))
poly_sf <- st_as_sf(poly_df, wkt = "geom")
ref_df <- do.call(rbind, lapply(names(demo), function(nm) {
  g  <- demo[[nm]]
  fr <- local_frame(g)
  cen   <- st_centroid(st_sfc(g))
  r_ref <- sqrt(as.numeric(st_area(st_sfc(g))) / pi)
  data.frame(panel = panel_of(nm),
             geom = st_as_text(normalize_geom(st_buffer(cen, dist = r_ref, nQuadSegs = 60), fr)))
}))
ref_sf <- st_as_sf(ref_df, wkt = "geom")

ggplot() +
  geom_sf(data = poly_sf, fill = "grey90", color = "grey40") +
  geom_segment(data = segment_df, aes(x, y, xend = xend, yend = yend, color = len), linewidth = 0.4) +
  geom_sf(data = ref_sf, fill = NA, color = "firebrick", linewidth = 0.8, linetype = "dashed") +
  scale_color_gradient(low = "steelblue", high = "firebrick", name = "segment length") +
  facet_wrap(~ panel) +
  coord_sf() +
  theme_gallery
```

Each panel's own equal-area disk reference (dashed) shows the same story a different way: the square nearly fills its own reference circle, the elongated rectangle spills far outside it, and the notched star sits in between - visually the same ranking the segment colours already suggest.

Look closely at the star panel and some segments visibly cut across a notch, leaving the polygon briefly before re-entering it. That's not a rendering artefact: span index only requires $s$ and $t$ themselves to be interior, never the line between them - unlike `convexity_index()`, which is built specifically around whether that connecting line stays inside (`vignette("c-understanding-convexity-index")`, and see the Introduction above for what this index has in common with the wider compactness literature). "Where this fits" below compares the two indices directly.

**Weighting** changes what this figure would show without changing what it means: points would be sampled in proportion to `weight` rather than uniform area, so segments would cluster wherever the density is highest rather than spreading evenly across the shape, and the dashed reference circle becomes the density-varying disk from `vignette("d-understanding-moment-of-inertia-index")`'s "Weighted version" (concentric rings, densest at the centre) rather than a flat one - the same reference construction "Weighted reference: the concentric annulus" below derives in full.

## Why not squared distance?

Squaring first would make the actual value trivial to compute - but at the cost of it being a different index in name only. For $s, t$ independent with density $\rho/W$,

$$\mathbb{E}|s-t|^2 = 2\,\text{Var}(s) = \frac{2J}{W}$$

where $J = \int_P \rho(s)\,|s-G|^2\,ds$ is exactly the polar moment of inertia `moment_of_inertia_index()` already computes about the mass centroid $G$. Both the disk reference ($J_{\text{ref}} = A^2/2\pi$) and the concentric-ring reference for a weighted profile are *also* already exact closed forms in that function. So a squared-distance version of span index would reduce to

$$\frac{D_{\text{ref}}^2}{D(\rho)^2} = \frac{2J_{\text{ref}}/W}{2J_{\text{actual}}/W} = \frac{J_{\text{ref}}}{J_{\text{actual}}}$$

which is exactly `moment_of_inertia_index()`'s own return value - the $W$'s cancel exactly. Plain (non-squared) distance has no such shortcut: by Jensen's inequality, $\mathbb{E}|s-t| \le \sqrt{\mathbb{E}|s-t|^2}$ strictly (equality would need $|s-t|$ almost-surely constant, never true on a continuous 2D region), and the gap depends on the *shape* of the distance distribution, not just its second moment. Concretely, the ratio $D/\sqrt{2J/W}$ is about $0.905$ for a disk but only about $0.816$ for a degenerate line segment - not a universal constant, so $D_{\text{ref}}/D_{\text{actual}} \ne \sqrt{J_{\text{ref}}/J_{\text{actual}}}$ in general. Span index is measuring something moment of inertia genuinely cannot see.

## The reference shapes really are optimal

Both reference shapes need to be true minimisers of $D$, not just plausible-looking candidates, for "index $\le 1$" to hold in general. The relevant tool is **Riesz's rearrangement inequality**: for fixed total mass, $\iint \rho(s)\rho(t)\,h(s-t)\,ds\,dt$ is maximised, among all spatial rearrangements of $\rho$, when $\rho$ is arranged symmetric-decreasing about a common centre - *provided* $h$ is itself symmetric-decreasing. Our kernel $|s-t|$ is *increasing*, not decreasing, but subtracting it from a large constant $C$ flips the sign without changing which arrangement is optimal (rearrangement preserves total mass, so the constant term is fixed): minimising $\iint \rho(s)\rho(t)|s-t|\,ds\,dt$ is the same problem as maximising $\iint \rho(s)\rho(t)(C - |s-t|)\,ds\,dt$, and $C-|s-t|$ *is* symmetric-decreasing.

- **Unweighted**: $\rho$ is an indicator function; its symmetric-decreasing rearrangement is, by definition, the indicator of the equal-area disk. Riesz confirms the disk is the exact global minimiser of $D$ among *every* measurable shape of that area, not just convex ones.
- **Weighted**: $\rho$'s rearrangement class is the same multiset of (weight, area) pairs the triangulation produced; its symmetric-decreasing rearrangement is concentric rings, sorted densest-to-centre - the same construction `moment_of_inertia_index()` already uses for its own weighted reference.

For the discrete (CDT-triangle) mesh this also has a direct finite proof by exchange: swapping the ring positions of two triangles where the denser one sits farther out strictly decreases the reference sum, since (as derived below) the pairwise kernel is increasing in each radius. Repeating every improving swap converges to the sorted order.

## Unweighted reference: the disk

The mean distance between two uniform random points in a disk of radius $R$ is a classical closed form, $\mathbb{E}|s-t| = \frac{128}{45\pi}R \approx 0.9054\,R$. With $R = \sqrt{A/\pi}$,

$$D_{\text{circle}}(A) = \frac{128\sqrt{A}}{45\,\pi^{3/2}}$$

## Weighted reference: the concentric annulus

No closed form exists ring-to-ring: unlike moment of inertia's $r^2$ kernel (which has a clean $r^4$ antiderivative), $|s-t|$ has none. But averaging over just the *angle* between two points at radii $r_1, r_2$ does:

$$g(r_1, r_2) = \frac{1}{2\pi}\int_0^{2\pi} \sqrt{r_1^2 + r_2^2 - 2 r_1 r_2 \cos\theta}\;d\theta = \frac{2}{\pi}(r_1+r_2)\,E(k), \qquad k = \frac{2\sqrt{r_1 r_2}}{r_1+r_2}$$

where $E$ is the complete elliptic integral of the second kind. Two sanity checks confirm it: $r_1=r_2=r$ gives $k=1$, $E(1)=1$, $g = 4r/\pi$ - the classical mean chord length between two random points on one circle. $r_2=0$ gives $k=0$, $E(0)=\pi/2$, $g=r_1$ - the (obviously correct) distance from the centre to any point on a circle of radius $r_1$.

```{r ellip-check}
k_vals <- c(0, 0.3, 0.5, 0.7, 1/sqrt(2), 0.9, 0.999, 1)
check <- vapply(k_vals, function(k) {
  ref <- integrate(function(phi) sqrt(1 - k^2 * sin(phi)^2), 0, pi/2)$value
  got <- shapeindices:::.ellip_E(k)
  abs(ref - got)
}, numeric(1))
```

(`shapeindices`'s internal `E(k)` is computed via the AGM algorithm, not a package dependency; cross-checked here against direct numerical integration of $E$'s own definition across `r length(k_vals)` values of $k$, largest discrepancy `r formatC(max(check), format = "e", digits = 1)`.)

This reduces the weighted reference to a 2D numerical integral - Gauss-Legendre quadrature on each ring's own radial marginal $f(r) = 2r/(r_{\text{hi}}^2 - r_{\text{lo}}^2)$ - summed over every pair of rings (sorted by density, same construction as above). One ring per triangle would make this pairwise sum $O(n_{\text{triangles}}^2)$ - fine at the mesh sizes this vignette uses, but on a real-world mesh of thousands of triangles the quadratic blow-up in time and memory can dwarf the rest of the computation. `span_index()` therefore caps the ring count (100 by default) and builds rings via a greedy pass bounded by density ratio, not triangle count: a ring closes once its own top-to-bottom density ratio would exceed 4x, or it's used its fair share of whatever triangles and ring budget remain, recomputed after every close. The density-ratio bound is what keeps the merged reference trustworthy: an outlier triangle - however extreme its weight relative to its neighbours - gets isolated in its own ring rather than averaged into one far wider than it should occupy, which could otherwise push the approximated reference above the true value and the index above 1. Below the ring cap, every triangle keeps its own ring exactly as derived above - no approximation at all.

A $3\times3$ grid of unit cells, weighted so the centre cell is heaviest and weight falls off toward the corners, shows the sorted-ring construction concretely - the same grid used in `vignette("a-basic-usage")`'s hole-punching example, reused here for its weight profile rather than to punch a hole.

```{r annulus-example, fig.width=9, fig.height=4.5}
cell <- function(cx, cy) st_polygon(list(rbind(
  c(cx-0.5, cy-0.5), c(cx+0.5, cy-0.5), c(cx+0.5, cy+0.5), c(cx-0.5, cy+0.5), c(cx-0.5, cy-0.5))))
ctr <- expand.grid(x = -1:1, y = -1:1)
w9  <- 10 - 3 * sqrt(ctr$x^2 + ctr$y^2)   # heaviest at centre, lightest at corners
grid9 <- st_sf(pop = w9, geometry = st_sfc(mapply(cell, ctr$x, ctr$y, SIMPLIFY = FALSE), crs = 3857))
poly9 <- st_union(st_geometry(grid9))
prep9 <- prepare_polygon(poly9)

# per-triangle weight: area-weighted overlay of each triangle against the
# grid cell weights, built by hand here since this demo triangulates the
# union directly via prepare_polygon() rather than going through
# shape_indices_sf(byrow = FALSE) itself (which, given real per-row
# weights like this, builds its mesh so no triangle can straddle a cell
# boundary in the first place, making this overlay step unnecessary)
ov <- st_intersection(st_sf(geometry = st_geometry(prep9$tri), tri_id = seq_len(nrow(prep9$tri))),
                       grid9)
ov$a <- as.numeric(st_area(st_geometry(ov)))
tri_w <- sapply(seq_len(nrow(prep9$tri)), function(i) {
  rows <- ov[ov$tri_id == i, ]
  if (nrow(rows) == 0) return(0)
  sum(rows$pop * rows$a) / sum(rows$a)
})

res9 <- span_index(poly9, prep = prep9, weight = tri_w)

# the reference geometry, purely for illustration (span_index() itself only
# returns the numeric D_ref, not this geometry): group by each grid CELL's
# own weight, not by the incidental CDT triangle count downstream (the CDT
# retriangulates the unioned 3x3 square from scratch once the internal cell
# boundaries are gone, so its triangle count has nothing to do with the 9
# original cells) - cells tied on weight are the same "parcel" for a sorted-
# ring arrangement, so this grid's 9 cells collapse to 3 rings (centre, the
# 4 edge-adjacent cells, the 4 corner cells), always fewer than 9
grid9$area <- as.numeric(st_area(st_geometry(grid9)))
tiers <- aggregate(area ~ pop, data = st_drop_geometry(grid9), sum)
tiers <- tiers[order(-tiers$pop), ]
r_hi <- sqrt(cumsum(tiers$area) / pi)
r_lo <- c(0, r_hi[-length(r_hi)])
cen9 <- st_centroid(poly9)
rings <- lapply(seq_along(r_hi), function(i) {
  outer_c <- st_buffer(cen9, r_hi[i], nQuadSegs = 60)
  if (r_lo[i] == 0) return(outer_c)
  st_difference(outer_c, st_buffer(cen9, r_lo[i], nQuadSegs = 60))
})
ring_sf <- st_sf(density = tiers$pop, geometry = st_sfc(lapply(rings, function(g) st_geometry(g)[[1]]), crs = st_crs(poly9)))

panel_a <- sprintf("Actual: weight by cell (span_index = %.3f)", res9$index)
panel_b <- sprintf("Reference: %d concentric rings", nrow(tiers))
combined <- rbind(
  st_sf(panel = panel_a, value = grid9$pop, geometry = st_geometry(grid9)),
  st_sf(panel = panel_b, value = ring_sf$density, geometry = st_geometry(ring_sf))
)
combined$panel <- factor(combined$panel, levels = c(panel_a, panel_b))

ggplot(combined) +
  geom_sf(aes(fill = value), color = "white") +
  scale_fill_gradient(low = "grey95", high = "firebrick") +
  facet_wrap(~ panel) +
  labs(fill = "weight/density") +
  theme_gallery
```

Nine cells collapse to `r nrow(tiers)` rings because only `r nrow(tiers)` distinct weight values exist among them (the centre cell, the 4 edge-adjacent cells, and the 4 corner cells each share a weight) - a sorted-ring reference always has *at most* as many rings as the mesh has distinct density levels, never more. The actual grid already has its heaviest cell at the centre, so its span index is close to (though not exactly) 1 - the residual gap is the grid's square, blocky footprint versus the reference's smooth circular rings, not a difference in *where* the mass sits. The real computation behind `res9` still triangulates and works ring-by-triangle as derived above (`r nrow(prep9$tri)` CDT triangles here, capped at 100 in general) - only this illustration's own reference panel is simplified down to the grid's `r nrow(tiers)` genuinely distinct density tiers, since a triangle-by-triangle breakdown would just be reproducing the incidental mesh resolution rather than the actual weighted-parcel structure the derivation above is about.

## Computing the actual value

The actual $D(\rho)$ is computed the same way `convexity_index()` computes its own deterministic mode - CDT the polygon, place quadrature points in each triangle - but simpler, since there's no line-clipping at all, just Euclidean distances between quadrature points. Cross-triangle pairs (different triangles) are well approximated this way with only a few points per triangle, since it's averaging over two whole separate regions.

**Same-triangle pairs are not.** With only 1 or 3 fixed quadrature points, most or all same-triangle "pairs" coincide (contributing zero distance), badly understating the true within-triangle spread - and this isn't negligible at the mesh sizes this package actually uses (tens to low hundreds of triangles). There's no closed form for a general triangle's own mean internal distance, so `span_index()` instead recursively splits each triangle by medial subdivision (connecting edge midpoints into 4 similar sub-triangles) and averages distances among the sub-triangle centroids - converging geometrically as subdivision deepens:

```{r self-term-convergence, echo=TRUE}
subdivide <- function(v) {
  m_ab <- (v[1,]+v[2,])/2; m_bc <- (v[2,]+v[3,])/2; m_ca <- (v[3,]+v[1,])/2
  list(rbind(v[1,],m_ab,m_ca), rbind(v[2,],m_bc,m_ab), rbind(v[3,],m_ca,m_bc), rbind(m_ab,m_bc,m_ca))
}
centroid_self_D <- function(v, k) {
  tris <- list(v)
  for (i in seq_len(k)) tris <- unlist(lapply(tris, subdivide), recursive = FALSE)
  cen <- t(vapply(tris, colMeans, numeric(2)))
  mean(as.matrix(dist(cen)))
}
v <- rbind(c(0,0), c(4,0), c(0,3))   # an arbitrary right triangle, legs 4 and 3
set.seed(1)
n_mc <- 300000
r1 <- runif(n_mc); r2 <- runif(n_mc); flip <- (r1+r2) > 1
r1[flip] <- 1-r1[flip]; r2[flip] <- 1-r2[flip]
P1 <- cbind(v[1,1]+r1*(v[2,1]-v[1,1])+r2*(v[3,1]-v[1,1]), v[1,2]+r1*(v[2,2]-v[1,2])+r2*(v[3,2]-v[1,2]))
r1 <- runif(n_mc); r2 <- runif(n_mc); flip <- (r1+r2) > 1
r1[flip] <- 1-r1[flip]; r2[flip] <- 1-r2[flip]
P2 <- cbind(v[1,1]+r1*(v[2,1]-v[1,1])+r2*(v[3,1]-v[1,1]), v[1,2]+r1*(v[2,2]-v[1,2])+r2*(v[3,2]-v[1,2]))
D_true <- mean(sqrt(rowSums((P1-P2)^2)))

convergence <- data.frame(
  depth = 0:6,
  n_centroids = 4^(0:6),
  D = vapply(0:6, function(k) centroid_self_D(v, k), numeric(1))
)
convergence$rel_error_pct <- 100 * (convergence$D - D_true) / D_true
knitr::kable(format = "html", convergence, digits = c(0, 0, 4, 3),
             col.names = c("subdivision depth", "centroids", "D estimate", "rel. error (%)"))
```

`span_index()` caps depth at 4 (256 centroids), but doesn't spend it on every triangle - the depth is area-adaptive, the same idea `radial_concentration_index()` uses for its own point cloud (see `vignette("f-understanding-radial-concentration-index")`), anchored so the mesh's own largest triangle still gets the full 256 centroids and smaller ones get proportionally fewer. Adaptivity matters more here than for `radial_concentration_index()`: the self-term needs the *full pairwise distance matrix* among one triangle's own centroids, an $O((4^{\text{depth}})^2)$ cost per triangle rather than $O(4^{\text{depth}})$, so quartering a small triangle's depth cuts its share of this cost sixteenfold, not just fourfold. On real Census-block-derived meshes with hundreds of triangles and large size variation, this cuts total self-term point count by 40-60x with the resulting index changing by well under 0.1% - invisible at the 3 decimal places indices are typically read to. One subtlety weighting introduces: depth adapts to each triangle's own *contribution* to the self-term sum ($w_i^2\sqrt{\text{area}_i}$, since a triangle's self-distance scales with its own linear size), not to raw area alone - a physically tiny triangle carrying almost all the weight still gets real depth, rather than having its self-distance collapse to exactly 0 the way pure area-based adaptation would give it. This is the same principle behind the reference's ratio-bounded rings above: weight can make triangle size a misleading proxy for how much a triangle actually matters to the sum.

**Monte Carlo mode** (`deterministic = FALSE`) sidesteps all of this: sample `n_lines` random point pairs directly (weighted sampling reuses the same per-triangle weighted sampler `convexity_index()` uses) and average their distance - no quadrature-cloud or subdivision needed, since there's no self/cross split to approximate in the first place. The argument is called `n_lines`, matching `convexity_index()`'s own Monte Carlo argument, even though no line geometry is actually built here - a line's length is exactly the distance between its two endpoints, so the name still fits, and sharing it means `shape_indices()` can set every mesh-based index's Monte Carlo sample count with one argument.

# Illustrations

## Basic shapes

```{r basic-shapes-table}
rows <- list(
  list(name = "square", geom = square),
  list(name = "disk (reference shape itself)", geom = disk)
)
for (nm in names(rectangles)) rows[[length(rows)+1]] <- list(name = paste("rectangle", nm), geom = rectangles[[nm]])
for (nm in names(stars_r))    rows[[length(rows)+1]] <- list(name = paste("star", nm), geom = stars_r[[nm]])

basic_tbl <- do.call(rbind, lapply(rows, function(r) {
  res <- span_index(r$geom)
  data.frame(shape = shape_thumb(r$geom), name = r$name,
             span_index = res$index, D = res$D, D_ref = res$D_ref)
}))
knitr::kable(format = "html", basic_tbl, digits = 3, row.names = FALSE, escape = FALSE)
```

The disk itself scores just above 1 rather than exactly 1 - the small, expected residual quadrature/subdivision error discussed above, not a modelling error (it shrinks further with a finer mesh or higher subdivision depth). Elongating a fixed-area rectangle drives the index down steadily; deepening a star's notches does too, though less sharply, since span index responds to *dispersal*, not boundary crossings specifically - a shape can score low here for being spread out even where every interior line stays inside it (`convexity_index() = 1`).

## Deterministic vs. Monte Carlo agreement

```{r deterministic-vs-mc}
shapes_for_mc <- list(square = square, `star (deep notch)` = stars_r[[5]])
mc_tbl <- do.call(rbind, lapply(names(shapes_for_mc), function(nm) {
  g <- shapes_for_mc[[nm]]
  det <- span_index(g)
  mc  <- suppressWarnings(span_index(g, deterministic = FALSE, n_lines = 20000, seed = 1))
  data.frame(shape = shape_thumb(g), name = nm,
             D_deterministic = det$D, D_monte_carlo = mc$D,
             index_deterministic = det$index, index_monte_carlo = mc$index)
}))
knitr::kable(format = "html", mc_tbl, digits = 4, row.names = FALSE, escape = FALSE)
```

The two methods agree to within about 1%, as expected for 20000 random point pairs.

## Things to watch out for: almost, but not quite, radius-only

`moment_of_inertia_index()` and `radial_concentration_index()` are both *exactly* blind to how mass is arranged angularly at a fixed radius, provided the arrangement keeps enough rotational symmetry to pin their reference point (centroid or median) in place - see their own "Things to watch out for" subsections in `vignette("d-understanding-moment-of-inertia-index")` and `vignette("f-understanding-radial-concentration-index")`. Span index shares their setup but not their conclusion, and the reason is structural: both siblings measure distance from a single fixed point, which depends only on radius, while span index measures distance *between two random points*, and $|s-t|^2 = r_s^2 + r_t^2 - 2r_sr_t\cos(\theta_s - \theta_t)$ depends on the angle *between* them too. Cutting a shape's mass into more, thinner angularly-separated pieces changes the distribution of that angular gap, so span index should move, even when its two siblings provably can't.

```{r blind-spot, echo=TRUE}
tbl_spokes <- do.call(rbind, lapply(names(spoke_shapes), function(nm) {
  poly <- st_sfc(spoke_shapes[[nm]])
  data.frame(shape = shape_thumb(spoke_shapes[[nm]]), name = nm,
             span = span_index(poly)$index,
             moment_of_inertia = moment_of_inertia_index(poly)$index,
             convexity = suppressWarnings(convexity_index(poly, deterministic = FALSE,
                                                            n_lines = 5000, seed = 1)$index))
}))
knitr::kable(format = "html", tbl_spokes, digits = 4, row.names = FALSE, escape = FALSE)
```

It does move - but only a little. Going from 2 broad wedges to 16 thin ones (the same total angular width, just cut finer) shifts `span` by about `r sprintf("%.1f%%", 100 * abs(tbl_spokes$span[1] - tbl_spokes$span[4]) / tbl_spokes$span[1])`, while `moment_of_inertia` doesn't move at all and `convexity` falls by over `r sprintf("%.0f%%", 100 * (tbl_spokes$convexity[1] - tbl_spokes$convexity[4]) / tbl_spokes$convexity[1])`. Span index is measurably *not* blind to this kind of rearrangement, unlike its two siblings - but it is far less sensitive to it than a shape's visibly widening gaps would suggest, because most of a segment's length is set by the two points' radii, with the angle between them only a second-order correction.

# Where this fits

`span_index()` is the no-fixed-reference-point member of a family of three. Its siblings measure the same "how spread out is the mass" question by anchoring to a single point: `moment_of_inertia_index()` uses *squared* distance to the centroid (`vignette("d-understanding-moment-of-inertia-index")`), and `radial_concentration_index()` uses plain distance to the geometric median (`vignette("f-understanding-radial-concentration-index")`). Span index needs no reference point at all - both ends of its distance are random - which is also why, alone among the three, it is not exactly blind to how mass is arranged angularly at a fixed radius (see "Things to watch out for" above): distance between two random points depends on the angle between them, not just their individual distances from any one centre.

`convexity_index()` (`vignette("c-understanding-convexity-index")`) sits outside that family entirely: it answers whether the segment connecting two interior points stays inside the shape, something span index was never trying to measure (see the Introduction above for how that relates to the wider compactness literature). The figure below plots span index against all three.

```{r glyph-helper}
# a small polygon, scaled to fit within `size` data-units and centred at
# (cx, cy) - used to draw each shape as its own outline directly in data
# space rather than a plain point, so glyphs work with facet_wrap() for
# free (it's just an ordinary geom_polygon() layer, no image/raster
# dependency needed)
glyph_polygon <- function(geom, cx, cy, size) {
  co  <- st_coordinates(st_sfc(geom))[, 1:2]
  rng <- max(diff(range(co[, 1])), diff(range(co[, 2])))
  data.frame(x = (co[, 1] - mean(range(co[, 1]))) / rng * size + cx,
             y = (co[, 2] - mean(range(co[, 2]))) / rng * size + cy)
}
```

```{r other-indices-comparison, fig.width=13, fig.height=4.5}
comparison_indices <- list(
  moment_of_inertia_index = moment_of_inertia_index,
  radial_concentration_index = radial_concentration_index,
  convexity_index = function(g) suppressWarnings(convexity_index(g, deterministic = FALSE, n_lines = 5000, seed = 1))
)
cmp_summary <- do.call(rbind, lapply(rows, function(r) {
  data.frame(name = r$name, span = span_index(r$geom)$index,
             moment_of_inertia_index = moment_of_inertia_index(r$geom)$index,
             radial_concentration_index = radial_concentration_index(r$geom)$index,
             convexity_index = comparison_indices$convexity_index(r$geom)$index)
}))
glyph_df <- do.call(rbind, lapply(rows, function(r) {
  y <- span_index(r$geom)$index
  do.call(rbind, lapply(names(comparison_indices), function(idx_name) {
    x <- comparison_indices[[idx_name]](r$geom)$index
    g <- glyph_polygon(r$geom, x, y, size = 0.045)
    g$panel <- idx_name
    g$id <- paste(r$name, idx_name)
    g
  }))
}))

ggplot(glyph_df) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey60") +
  geom_polygon(aes(x, y, group = id), fill = "steelblue", color = "grey20", linewidth = 0.3) +
  facet_wrap(~ panel) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = NULL, y = "span_index")
```

The `convexity_index()` panel, leftmost, looks different in kind, not just degree: every rectangle and the square sit in a single vertical line at convexity 1, however elongated, since a straight line between two points of a convex shape never has to leave it - only the stars move off that line, both indices falling together as notches deepen, because part of span index's own drop there is detecting the same re-entrancy convexity index measures directly.

The two "distance from a centre" siblings, by contrast, agree with span index in ranking throughout - elongating rectangles and deepening star notches drive all three down together - but their own two panels are not equally close to the dashed $y=x$ line. Span index tracks `radial_concentration_index()` tightly (mean absolute gap `r sprintf("%.3f", mean(abs(cmp_summary$span - cmp_summary$radial_concentration_index)))`) but sits much further from `moment_of_inertia_index()` (mean absolute gap `r sprintf("%.3f", mean(abs(cmp_summary$span - cmp_summary$moment_of_inertia_index)))`, about `r sprintf("%.0f", mean(abs(cmp_summary$span - cmp_summary$moment_of_inertia_index)) / mean(abs(cmp_summary$span - cmp_summary$radial_concentration_index)))`x larger) - the same *plain vs. squared* distinction from "Why not squared distance?" above, showing up again here rather than just in the derivation: `radial_concentration_index()` and span index both use plain distance (to a point, and between two points, respectively), while `moment_of_inertia_index()` squares it. The elongated rectangles drift furthest from the line against `moment_of_inertia_index()` specifically, matching that same mechanism - squared distance penalises their far-flung mass harder than plain distance does. Neither panel sits exactly on the line, so span index isn't a rescaling of either sibling, but on shapes like these it is close enough to `radial_concentration_index()` to track it almost one-for-one.

# Key takeaways

- `span_index()` compares $D$ (mean plain distance between two independent random interior points) to the same quantity for an equal-area disk, or - weighted - a concentric annulus, both provably optimal via Riesz's rearrangement inequality. 1 exactly when the shape matches its reference.
- It is a genuinely different quantity from `moment_of_inertia_index()`, not a rescaling of it: squaring the distance first would collapse span index exactly onto moment of inertia (the $W$'s cancel), but plain distance has no such shortcut, by Jensen's inequality.
- Neither reference has a fully closed form once weighted: the disk case does, but the concentric-annulus case needs the complete elliptic integral of the second kind, evaluated ring-pair by ring-pair.
- Getting the *actual* value right needs more than the deterministic quadrature `convexity_index()` uses: same-triangle point pairs need area-adaptive medial subdivision, not a fixed few quadrature points, or the within-triangle spread is badly understated.
- It is almost, but not quite, blind to how mass is arranged angularly at a fixed radius - unlike `moment_of_inertia_index()`/`radial_concentration_index()`, which are blind to it exactly. Distance between two random points depends on the angle between them, so span index does move as a shape's mass is cut into more, thinner angularly-separated pieces, just far less than the widening gaps might suggest (see "Things to watch out for" above).
- `deterministic = FALSE` (Monte Carlo) agrees with the default deterministic mode to within about 1% at realistic sample sizes, with no quadrature or subdivision approximation of its own.
- $s$ and $t$ only need to be interior themselves - the segment between them can leave the shape and come back, unlike `convexity_index()`, which is built specifically to measure that (Introduction). Not an oversight: on convex shapes the two indices answer fully orthogonal questions (every rectangle sits on one vertical line of `convexity_index()` = 1 regardless of elongation), but on concave shapes part of span index's own drop is detecting the same re-entrancy `convexity_index()` measures directly (see "Where this fits" above).
