---
title: "7. Understanding Directional Balance Index"
author: "Nikhil Kaza"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{7. Understanding Directional Balance Index}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
  width: auto !important;
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.align = "center", cache = TRUE)
```

```{r load, cache=FALSE, echo=TRUE}
#| code-fold: true
library(shapeindices)
library(sf)
library(ggplot2)

theme_set(theme_minimal(base_size = 11))
theme_gallery <- theme_void(base_size = 10) +
  theme(strip.text = element_text(size = 9, face = "bold"))
```

```{r shapes, echo=TRUE}
#| code-fold: true
square <- st_polygon(list(rbind(c(0,0), c(10,0), c(10,10), c(0,10), c(0,0))))

make_regular_ngon <- function(n, r = 1, center = c(0, 0)) {
  ang <- seq(0, 2*pi, length.out = n + 1)[1:n]
  st_polygon(list(rbind(cbind(center[1] + r*cos(ang), center[2] + r*sin(ang)), center + c(r, 0))))
}
hexagon <- make_regular_ngon(6, 5)
disk <- st_buffer(st_sfc(st_point(c(0, 0))), dist = 5.64, nQuadSegs = 60)[[1]]

make_star <- function(n_points, r_outer = 1, r_inner = 0.5, center = c(0, 0)) {
  n <- n_points * 2
  angles <- seq(pi/2, pi/2 + 2*pi, length.out = n + 1)[1:n]
  radii  <- rep(c(r_outer, r_inner), n_points)
  x <- center[1] + radii * cos(angles); y <- center[2] + radii * sin(angles)
  st_polygon(list(rbind(cbind(x, y), c(x[1], y[1]))))
}
star6 <- make_star(6, 5, 2.5)

make_dumbbell <- function(r_top, r_bottom, cy = 10) {
  top    <- st_buffer(st_sfc(st_point(c(0, cy))), r_top, nQuadSegs = 40)
  bottom <- st_buffer(st_sfc(st_point(c(0, -cy))), r_bottom, nQuadSegs = 40)
  neck   <- st_polygon(list(rbind(c(-0.3,-cy), c(0.3,-cy), c(0.3,cy), c(-0.3,cy), c(-0.3,-cy))))
  st_make_valid(st_union(st_sfc(c(top[[1]], bottom[[1]], neck))))[[1]]
}
dumbbell_sym  <- make_dumbbell(3, 3)     # equal lobes
dumbbell_asym <- make_dumbbell(4, 1.2)   # unequal lobes
```

# Introduction

`directional_balance_index()` measures whether a (multi)polygon's own mass is spread evenly across every bearing from its centroid, or leans toward some bearings more than others. Concretely: $R$ is the magnitude of the mean resultant vector of $e^{i\theta(x)}$, where $\theta(x)$ is the angle of point $x$ as seen from the shape's own mass centroid $G$ - and the index is $1 - R$, in $[0, 1]$.

**What this index uses - and what it deliberately ignores.** Unlike `moment_of_inertia_index()` and `moment_isotropy_index()`, which integrate position *squared*, this index uses only the **bearing** $\theta(s)$, never the distance. Distance would carry no information here anyway: $G$ is the mass centroid by definition, so the raw first moment of position, $\int_P \rho(s)(s - G)\,ds$, is identically zero for *every* shape. $e^{i\theta(s)}$ discards distance and keeps only direction, and that is precisely what makes the quantity informative - it asks whether the shape's own area, bearing by bearing, is spread evenly around the centroid or piled up toward some directions. One reading to guard against: **balanced does not mean uniform**. $R = 0$ says the directional pulls *cancel*, not that mass is spread evenly across every bearing - "Things to watch out for" in the Illustrations section below shows exactly a shape where that distinction matters.

```{r glance-helper}
# a small base64-embedded PNG of a shape, for use as a table cell in place
# of a text label - knitr::image_uri() does the embedding, so no new
# package dependency (it's already pulled in transitively via knitr, which
# every vignette in this package already Suggests)
shape_thumb <- function(geom, size_px = 40, col = "grey75") {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(st_geometry(st_sfc(geom)), col = col, border = "grey30")
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}

# same idea, but shaded by a triangle mesh's own per-triangle weight instead
# of a flat fill - for tables comparing weight DISTRIBUTIONS on one fixed
# shape rather than comparing different shapes
weight_thumb <- function(tri, w, size_px = 40) {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  pal  <- grDevices::colorRampPalette(c("grey95", "firebrick"))(100)
  cols <- pal[as.integer(cut(w, 100, labels = FALSE))]
  plot(st_geometry(tri), col = cols, border = NA)
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}
```

# Deriving the index

## The resultant vector

For density $\rho$ over the polygon, a point $s = (x, y) \in P$ with coordinates $x, y$, and mass centroid $G = (G_x, G_y)$, define $\theta(s) = \operatorname{atan2}(y - G_y,\ x - G_x)$ - the bearing of $s$ as seen from $G$ - and

$$R = \left\lvert \frac{1}{W}\int_P \rho(s)\, e^{i\theta(s)}\,ds \right\rvert, \qquad W = \int_P \rho(s)\,ds$$

This is exactly the **mean resultant length** from circular/directional statistics (Rayleigh 1880's test for uniformity on a circle is built on the same quantity[^rayleigh]), here applied to a polygon's own interior angular mass distribution rather than to literal directional data like wind bearings. That specific application - bearings of interior area from a shape's own mass centroid, rather than literal directional observations - doesn't appear to have an established name in the shape-compactness literature the rest of this package draws from (see `vignette("d-understanding-span-index")`'s Introduction); treat it as a direct application of a classical tool to a new setting, not a rediscovery of a named index. `directional_balance_index()` is $1 - R$.

[^rayleigh]: Rayleigh, Lord (1880). On the resultant of a large number of vibrations of the same pitch and of arbitrary phase. *Philosophical Magazine*, 10(60), 73-78. Mardia, K. V., & Jupp, P. E. (2000). *Directional Statistics*. Wiley, is the standard modern reference for the mean resultant length and its uses.

Each interior point $s$ contributes a *unit* arrow pointing along its own bearing $\theta(s)$ from the centroid - distance discarded, direction kept. $R$ is the length of the average of all those unit arrows. When the bearings are spread evenly the arrows cancel and $R \approx 0$ (index $\approx 1$); when the mass leans one way they reinforce and $R$ grows toward 1 (index toward 0). The figure colours sample points by their bearing and draws the resulting average arrow from the centroid:

```{r resultant-vector, fig.width=9, fig.height=4.5}
set.seed(2)
resultant_panel <- function(g, dx = 0) {
  gt <- st_sfc(g) + c(dx, 0)
  db <- directional_balance_index(gt)
  G  <- st_coordinates(db$centroid)[1, 1:2]
  p  <- st_coordinates(st_sample(gt, size = 350, type = "random", exact = TRUE))[, 1:2]
  th <- atan2(p[, 2] - G[2], p[, 1] - G[1])
  list(pts = data.frame(x = p[, 1], y = p[, 2],
                        col = hcl(h = (th * 180 / pi) %% 360, c = 80, l = 65)),
       cen = data.frame(x = G[1], y = G[2]),
       R = db$R, angle = db$mean_angle,
       ring = { rc <- st_coordinates(gt)[, 1:2]; data.frame(x = rc[, 1], y = rc[, 2]) },
       label = sprintf("index %.3f", db$index), lx = G[1])
}
disk_p <- resultant_panel(disk, dx = 0)
db_p   <- resultant_panel(dumbbell_asym, dx = 16)
arrow_scale <- 6 / max(disk_p$R, db_p$R)
arrows <- do.call(rbind, lapply(list(disk_p, db_p), function(q) data.frame(
  x = q$cen$x, y = q$cen$y,
  xend = q$cen$x + arrow_scale * q$R * cos(q$angle),
  yend = q$cen$y + arrow_scale * q$R * sin(q$angle))))
rings  <- do.call(rbind, Map(function(q, s) transform(q$ring, shape = s), list(disk_p, db_p), c("a", "b")))
pts    <- do.call(rbind, lapply(list(disk_p, db_p), `[[`, "pts"))
cens   <- do.call(rbind, lapply(list(disk_p, db_p), `[[`, "cen"))
labs   <- data.frame(x = c(disk_p$lx, db_p$lx), y = 16,
                     lab = c(disk_p$label, db_p$label))

ggplot() +
  geom_polygon(data = rings, aes(x, y, group = shape), fill = NA, color = "grey55") +
  geom_point(data = pts, aes(x, y, color = col), size = 0.5) +
  scale_color_identity() +
  geom_segment(data = arrows, aes(x, y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.9, color = "black") +
  geom_point(data = cens, aes(x, y), shape = 3, size = 2.5, stroke = 1) +
  geom_text(data = labs, aes(x, y, label = lab), size = 3.4) +
  coord_equal() +
  theme_gallery
```

The disk's bearings fill every direction evenly, the unit arrows cancel, and the average arrow collapses to essentially nothing - index $\approx 1$. The lopsided dumbbell's mass leans toward its bigger (upper) lobe, so its arrow is long and points that way, dropping the index. The arrow's *direction* is `mean_angle`; its *length* (relative to a fully one-directional shape) is $R$.

## Why this is always in [0, 1]

$|R| \le 1$ by the triangle inequality on a complex-valued expectation: $\left|E[e^{i\theta}]\right| \le E\left|e^{i\theta}\right| = E[1] = 1$. That is the entire proof - no reference shape, no rearrangement inequality (the argument behind `convexity_index()`/`span_index()`/`radial_concentration_index()`'s bounds), no matrix fact (the argument behind `moment_isotropy_index()`'s). It also holds for *any* probability measure over bearings - continuous, or a finite weighted point cloud alike - so the discrete estimates both computation modes below produce respect the bound exactly, for any mesh and any sample size, not just approximately or in the limit.

## No closed form, unlike the moment family

$I_{xx}$/$I_{yy}$/$I_{xy}$ are polynomial in the vertex coordinates - dividing by radius to get a pure bearing breaks that structure, so $\int e^{i\theta(x)}\,dA$ over a triangle has no closed form. `directional_balance_index()` is genuinely approximate in both modes, the same situation `convexity_index()`/`span_index()`/`radial_concentration_index()` are in (not `moment_of_inertia_index()`/`moment_isotropy_index()`, which are exact regardless of mode). One practical consequence, visible directly:

```{r no-closed-form-demo, echo=TRUE}
directional_balance_index(st_sfc(square))$index   # vertex-exact symmetric shape
directional_balance_index(st_sfc(disk))$index     # curve approximated by 240 segments
```

The square scores exactly 1 (its mesh happens to be perfectly mirror-symmetric), while the segment-approximated disk lands a hair below. The gap is quadrature resolution, and refining the internal subdivision drives it to zero:

```{r discretization-convergence}
prep <- prepare_polygon(st_sfc(disk))
tri  <- prep$tri
w    <- tri$area
G    <- shapeindices:::.mass_centroid(st_geometry(tri), w)
depths <- c(0, 2, 4, 6, 8)
Rs <- vapply(depths, function(d) {
  cloud <- shapeindices:::.radial_point_cloud(tri, w, max_depth = d)
  shapeindices:::.resultant_from_cloud(cloud$p, cloud$w, G)$R
}, numeric(1))
data.frame(subdivision_depth = depths, R = Rs, index = 1 - Rs) |> knitr::kable(format = "html", digits = 6)
```

A disk approximated by straight boundary segments is, in the continuous limit, exactly balanced ($R = 0$) - but a CDT of a many-sided polygon isn't necessarily triangulated with perfect reflection symmetry, so a small residual survives at shallow subdivision depth. This is ordinary discretization error, and it shrinks predictably as depth increases. `directional_balance_index()` uses depth 4 by default (the same area-adaptive depth `radial_concentration_index()` uses), so expect a disk to score very close to, but not bit-for-bit exactly, 1 - anything within about $10^{-3}$ of 1 should be read as "balanced".

# Illustrations

## Basic shapes: directional balance vs. the property it isn't

```{r basic-table}
rows <- list(square = square, hexagon = hexagon, disk = disk, star = star6,
             symmetric_dumbbell = dumbbell_sym, asymmetric_dumbbell = dumbbell_asym)
tbl <- do.call(rbind, lapply(names(rows), function(nm) {
  poly <- st_sfc(rows[[nm]])
  data.frame(shape = shape_thumb(rows[[nm]]), name = nm,
             directional_balance = directional_balance_index(poly)$index,
             moment_isotropy = moment_isotropy_index(poly)$index,
             moment_of_inertia = moment_of_inertia_index(poly)$index)
}))
knitr::kable(format = "html", tbl, digits = 3, row.names = FALSE, escape = FALSE)
```

`moment_isotropy_index()` is a *second*-moment/axis-preference measure - does the mass prefer an axis, agnostic to which end of it - while `directional_balance_index()` is a *first*-harmonic/direction-preference measure - does the mass lean toward one bearing specifically. The symmetric dumbbell is the cleanest case where they diverge: low isotropy (it strongly prefers the north-south axis) but perfect directional balance (it doesn't prefer north *over* south, or vice versa). A shape "pointing north" and its mirror image "pointing south" would score identically on `moment_isotropy_index()` but oppositely on `directional_balance_index()`'s own `mean_angle` (not on `index` itself, which reports only magnitude).

## Weighting: free, for the same reason moment isotropy's is

No separate reference shape means no separate weighted derivation - the resultant vector is computed directly from whatever `weight` substitutes for triangle area, exactly like `moment_isotropy_index()`.

```{r weighted-dumbbell, echo=TRUE, fig.width=8, fig.height=4}
prep <- prepare_polygon(st_sfc(dumbbell_sym))
tri  <- prep$tri
cen  <- st_coordinates(st_centroid(st_geometry(tri)))
w_top <- ifelse(cen[, 2] > 0, 3, 1)  # 3x weight on the top lobe, still symmetric-shape but not symmetric-mass

plain    <- directional_balance_index(st_sfc(dumbbell_sym), prep = prep)
weighted <- directional_balance_index(st_sfc(dumbbell_sym), prep = prep, weight = w_top)

data.frame(
  weighting = c(weight_thumb(tri, rep(1, nrow(tri))), weight_thumb(tri, w_top)),
  name = c("uniform (area)", "3x weight on top lobe"),
  index = c(plain$index, weighted$index),
  mean_angle_deg = c(NA, weighted$mean_angle * 180 / pi)
) |> knitr::kable(format = "html", digits = c(NA, NA, 3, 1), escape = FALSE)
```

The shape's own boundary never changes here - a geometrically symmetric dumbbell - but concentrating mass in the top lobe alone breaks the angular cancellation, dropping the index and pointing `mean_angle` north, toward the heavier lobe.

## Things to watch out for

### Balanced doesn't mean uniform

```{r blind-spot, echo=TRUE}
res_sym  <- directional_balance_index(st_sfc(dumbbell_sym))
res_asym <- directional_balance_index(st_sfc(dumbbell_asym))
data.frame(
  shape = c(shape_thumb(dumbbell_sym), shape_thumb(dumbbell_asym)),
  name = c("symmetric dumbbell (equal lobes)", "asymmetric dumbbell (top lobe bigger)"),
  R = c(res_sym$R, res_asym$R),
  index = c(res_sym$index, res_asym$index),
  mean_angle_deg = c(NA, res_asym$mean_angle * 180 / pi)
) |> knitr::kable(format = "html", digits = c(NA, NA, 6, 4, 1), escape = FALSE)
```

The symmetric dumbbell scores `index = 1` - two equal masses on opposite bearings from $G$ cancel *exactly* in the complex sum, regardless of how badly dispersed the shape is along its own axis. Compare `moment_of_inertia_index()` and `span_index()`, both of which score this shape poorly (it's about as far from a disk as a shape this compact can get):

```{r blind-spot-compare, echo=TRUE}
data.frame(
  shape = shape_thumb(dumbbell_sym),
  name = "symmetric dumbbell",
  directional_balance = res_sym$index,
  moment_isotropy = moment_isotropy_index(st_sfc(dumbbell_sym))$index,
  moment_of_inertia = moment_of_inertia_index(st_sfc(dumbbell_sym))$index,
  span = span_index(st_sfc(dumbbell_sym))$index
) |> knitr::kable(format = "html", digits = 3, row.names = FALSE, escape = FALSE)
```

Making the two lobes unequal breaks the cancellation: the bigger lobe pulls more area to bear in its own direction from the (still centrally located) centroid, dropping the index to `r sprintf("%.3f", res_asym$index)` with `mean_angle` pointing toward it (the bigger lobe sits due north; `mean_angle` reads `r sprintf("%.1f", res_asym$mean_angle * 180/pi)` degrees, `atan2` convention).

**This is a feature, not purely a limitation.** Paired with `moment_of_inertia_index()`/`span_index()`/`moment_isotropy_index()` (all of which score the symmetric dumbbell badly), the two families together distinguish "symmetric two-sided reach" - fine here, bad on the others - from "one-sided reach" - bad on both. That distinction is plausibly relevant to redistricting analysis: a district extending outward in one direction only reads differently from one extending outward symmetrically in two, even though both can look equally "dispersed" by a single distance-based measure.

### Monte Carlo mode's small, structural bias

`deterministic = FALSE` draws points directly from the weighted density (the same mechanism `radial_concentration_index()` uses) and computes $\hat{R} = |\text{mean}(e^{i\theta_i})|$ over the sample, with $\theta_i$ measured relative to the *exact* centroid $G$ (found in closed form regardless of mode). Because $\hat{R}$ is the magnitude of a sample-mean vector - a nonlinear transform - Jensen's inequality means $E[\hat R] \ge |E[\text{sample mean})]| = R$: for a truly balanced shape, $\hat R$ is essentially never exactly 0 in a finite sample, so the Monte Carlo index sits systematically slightly *below* the true value. The gap shrinks as `n_lines` grows:

```{r mc-bias, echo=TRUE}
gap <- function(n) 1 - directional_balance_index(st_sfc(square), deterministic = FALSE,
                                                   n_lines = n, seed = 1)$index
ns <- c(200, 1000, 5000, 20000)
data.frame(n_lines = ns, gap_from_1 = vapply(ns, gap, numeric(1))) |> knitr::kable(format = "html", digits = 5)
```

The returned index carries this bias uncorrected, because it cannot be fixed on its own scale: **no unbiased estimator of $1 - R$ exists** for any finite sample. The obstruction is the kink of $R = |E[e^{i\theta}]|$ at exact balance - the same structural reason a sample standard deviation cannot be unbiased even though a sample variance can - and it sits at $R = 0$, exactly where the bias is worst, which is also where asymptotic corrections (whose leading term carries a $1/R$) diverge.

What *does* admit an exactly unbiased estimate is the same quantity on a squared scale, $1 - R^2$ - same $[0, 1]$ range, same direction, same shapes scoring exactly 1. Because the sampled bearings are i.i.d., $E[\cos(\theta_j - \theta_k)] = R^2$ for any two distinct draws, which makes

$$1 - \frac{n\,\hat{R}^2 - 1}{n - 1}$$

exactly unbiased for $1 - R^2$, for every weight distribution and every $n \ge 2$ - computable in one line from the returned `R` and your own `n_lines`. Two things to know if you use it: it can land slightly *above* 1 for near-balanced shapes - necessarily, since no estimator confined to $[0, 1]$ can be unbiased at the boundary - and truncating it back to 1 would reintroduce the bias, so leave it be. And since squaring compresses the top of the scale (real near-balanced shapes crowd into $[0.95, 1]$ on $1 - R^2$), treat it as a tool for unbiased estimation and batch averaging, not as a more readable index. If you just want a better point estimate of $1 - R$ itself, increase `n_lines` or use `deterministic = TRUE`, which carries no statistical bias at all.

# Where this fits

`directional_balance_index()` adds a genuinely new question to this package's family: not "is this shape convex" (`convexity_index()`, `hull_ratio_index()`), "how far is mass dispersed from the centre" (`moment_of_inertia_index()`, `span_index()`, `radial_concentration_index()`), "how round is the boundary/bounding shape" (`polsby_popper_index()`, `reock_index()`, `width_length_ratio_index()`), or "does the mass distribution prefer an axis" (`moment_isotropy_index()`) - but "does the mass distribution prefer a *bearing*". See `vignette("j-nc-counties-comparison")` for how it relates to the other nine on 100 real county boundaries.

# Key takeaways

- `directional_balance_index()` is $1 - R$, where $R$ is the magnitude of the mean resultant vector of unit bearings $e^{i\theta(s)}$ seen from the mass centroid - a first-harmonic, direction-preference measure, not a dispersal measure: distance from the centroid never enters the definition.
- It is bounded in $[0, 1]$ by the triangle inequality alone, the same elementary-fact style of proof `moment_isotropy_index()` uses, but no closed form exists for either computation mode - both `deterministic = TRUE` (subdivision-based quadrature) and `deterministic = FALSE` (Monte Carlo) are genuinely approximate, unlike the exact moment family.
- $R = 0$ means the directional pulls *cancel*, not that mass is spread evenly across every bearing: a symmetric two-lobed dumbbell scores a perfect 1 despite being about as far from a disk as a compact shape can get, because its two lobes pull in exactly opposite directions (see "Things to watch out for" above). Pair it with `moment_of_inertia_index()`/`span_index()`/`moment_isotropy_index()`, all of which score that same shape badly, to catch what it misses.
- Weighting is free, for the same reason `moment_isotropy_index()`'s is: there's no reference shape to re-derive, so a weighted index is just the resultant vector computed directly from whatever `weight` substitutes for triangle area.
- `deterministic = FALSE` carries a small, structural, uncorrectable-on-its-own-scale bias toward *higher* apparent imbalance, worst exactly at true balance ($R = 0$) - a consequence of $R$ being a magnitude, not an average, so Jensen's inequality bites. `deterministic = TRUE` (the default) carries no such bias; use `n_lines` in the tens of thousands if you need `deterministic = FALSE` to average out the gap instead.
- It asks a question none of this package's other nine indices ask: not how dispersed the mass is, or whether it prefers an axis, but whether it prefers a bearing - complementary to, not redundant with, `moment_isotropy_index()`'s axis-preference question.
