---
title: "7. Understanding Moment Isotropy Index"
author: "Nikhil Kaza"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    theme: flatly
vignette: >
  %\VignetteIndexEntry{7. Understanding Moment Isotropy Index}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

<style>
table {
  width: auto !important;
  max-width: 100%;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                       fig.align = "center", cache = TRUE)
```

```{r load, cache=FALSE, echo=TRUE}
#| code-fold: true
library(shapeindices)
library(sf)
library(ggplot2)

theme_set(theme_minimal(base_size = 11))
theme_gallery <- theme_void(base_size = 10) +
  theme(strip.text = element_text(size = 9, face = "bold"))
```

```{r shapes, echo=TRUE}
#| code-fold: true
square <- st_polygon(list(rbind(c(0,0), c(10,0), c(10,10), c(0,10), c(0,0))))
make_rect <- function(w, h) st_polygon(list(rbind(c(0,0), c(w,0), c(w,h), c(0,h), c(0,0))))
aspect_seq <- c(1, 2, 4, 10, 20)
rectangles <- lapply(aspect_seq, function(a) make_rect(sqrt(100 * a), sqrt(100 / a)))
names(rectangles) <- sprintf("aspect %gx", aspect_seq)

make_star <- function(n_points, r_outer = 1, r_inner = 0.5, center = c(0, 0)) {
  n <- n_points * 2
  angles <- seq(pi/2, pi/2 + 2*pi, length.out = n + 1)[1:n]
  radii  <- rep(c(r_outer, r_inner), n_points)
  x <- center[1] + radii * cos(angles); y <- center[2] + radii * sin(angles)
  st_polygon(list(rbind(cbind(x, y), c(x[1], y[1]))))
}
ratio_seq <- c(0.9, 0.7, 0.5, 0.3, 0.15)
stars_r <- lapply(ratio_seq, function(r) make_star(6, 5, 5 * r))
names(stars_r) <- sprintf("notch ratio %.2f", ratio_seq)

make_regular_ngon <- function(n, r = 1, center = c(0, 0)) {
  ang <- seq(0, 2*pi, length.out = n + 1)[1:n]
  st_polygon(list(rbind(cbind(center[1] + r*cos(ang), center[2] + r*sin(ang)), center + c(r, 0))))
}
hexagon <- make_regular_ngon(6, 5)

disk <- st_buffer(st_sfc(st_point(c(0, 0))), dist = 5.64, nQuadSegs = 60)[[1]]

# thin rectangular arms radiating from a small hub, with real empty space
# between them - unlike stars_r above (a single concave outline, no gaps
# of zero mass anywhere), a pinwheel has genuine angular voids, which is
# what makes the effect below visually undeniable rather than a curiosity
# about squares and hexagons
make_pinwheel <- function(n_arms, arm_len = 10, arm_width = 1, hub_r = 0.5) {
  angles <- seq(0, 2 * pi, length.out = n_arms + 1)[1:n_arms]
  hub <- st_buffer(st_sfc(st_point(c(0, 0))), hub_r, nQuadSegs = 30)[[1]]
  arms <- lapply(angles, function(a) {
    tip  <- c(arm_len * cos(a), arm_len * sin(a))
    perp <- c(-sin(a), cos(a)) * arm_width / 2
    st_polygon(list(rbind(perp, tip + perp, tip - perp, -perp, perp)))
  })
  Reduce(function(p, q) st_union(st_sfc(p), st_sfc(q))[[1]], c(list(hub), arms))
}
pinwheels <- lapply(2:6, make_pinwheel)
names(pinwheels) <- sprintf("%d arms", 2:6)
```

# Introduction

`moment_isotropy_index()` measures how anisotropic - how much it prefers one direction over others - a (multi)polygon's mass distribution is: the ratio of the smaller to the larger principal moment of its mass inertia tensor, in $(0, 1]$. It reuses `moment_of_inertia_index()`'s own machinery exactly - same triangulation, same mass centroid, same $I_{xx}$/$I_{yy}$ - plus one new quantity, the product of inertia $I_{xy}$, needed to find the tensor's eigenvalues.

**How this relates to classical eccentricity.** If you know ellipses, the natural first guess is that this index *is* eccentricity - it isn't, in two ways worth having clear before reading further. For an ellipse with semi-axes $a \ge b$, this index works out to $b^2/a^2 = 1 - e^2$, where $e = \sqrt{1-(b/a)^2}$ is the classical eccentricity: related, but not the same formula. And the two run in opposite directions - eccentricity *increases* from 0 (circle) to 1 (degenerate) as a shape elongates, while this index *decreases* from 1 toward 0, matching the 1-is-ideal convention every index in this package follows. "Isotropy" states the property directly: a high value means the mass distribution treats every direction alike.

It's also built differently from the package's reference-shape indices, and worth being explicit about that up front. `convexity_index()`, `moment_of_inertia_index()`, `span_index()`, and `radial_concentration_index()` are all "actual value vs. a provably optimal reference shape" comparisons - a rearrangement inequality or bathtub-principle argument shows a disk (or, weighted, concentric rings) is the unique best-possible arrangement, and the index is how close the real shape gets to that ceiling. `moment_isotropy_index()` has no reference shape at all: it compares the shape's own two principal moments to *each other*, an elementary fact about positive semi-definite matrices, not a rearrangement inequality. One consequence worth flagging immediately: an index of 1 does **not** mean "this is a disk" the way it does for `moment_of_inertia_index()`/`span_index()`/`radial_concentration_index()` - it means the mass distribution is rotationally isotropic about its own centroid, which a disk achieves, but so does a square, a regular hexagon, or anything with three-fold or higher rotational symmetry.

The ratio of an inertia tensor's smaller to larger eigenvalue, exactly this index's construction, is an established anisotropy measure - it comes from a different field than the geography-derived compactness literature the rest of this package's indices draw from (see `vignette("e-understanding-span-index")`'s Introduction). Materials science and image analysis use it under names like "isotropy index" or "elongation factor," most notably in the Minkowski-tensor anisotropy framework of Schröder-Turk et al. (2013).[^schroder-turk] That framework's own isotropy index, $\beta = \lambda_{\min}/\lambda_{\max}$, is the same eigenvalue ratio computed here, just applied to a mass inertia tensor rather than a Minkowski tensor.

[^schroder-turk]: Schröder-Turk, G. E., Mickel, W., Kapfer, S. C., Schaller, F. M., Breidenbach, B., Hug, D., & Mecke, K. (2013). Minkowski tensors of anisotropic spatial structure. *New Journal of Physics*, 15, 083028.

```{r glance-helper}
# a small base64-embedded PNG of a shape, for use as a table cell/header in
# place of a text label - knitr::image_uri() does the embedding, so no new
# package dependency (it's already pulled in transitively via knitr, which
# every vignette in this package already Suggests)
shape_thumb <- function(geom, size_px = 40, col = "grey75") {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(st_geometry(st_sfc(geom)), col = col, border = "grey30")
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}

# same idea, but shaded by a triangle mesh's own per-triangle weight instead
# of a flat fill - for tables comparing weight DISTRIBUTIONS on one fixed
# shape rather than comparing different shapes
weight_thumb <- function(tri, w, size_px = 40) {
  f <- tempfile(fileext = ".png")
  grDevices::png(f, width = size_px, height = size_px, bg = "transparent", res = 96)
  par(mar = c(0.5, 0.5, 0.5, 0.5))
  pal  <- grDevices::colorRampPalette(c("grey95", "firebrick"))(100)
  cols <- pal[as.integer(cut(w, 100, labels = FALSE))]
  plot(st_geometry(tri), col = cols, border = NA)
  grDevices::dev.off()
  uri <- knitr::image_uri(f)
  unlink(f)
  sprintf('<img src="%s" width="%d" height="%d" style="vertical-align:middle"/>', uri, size_px, size_px)
}
```

# Deriving the index

## The mass inertia tensor

For density $\rho$ (piecewise-constant across the CDT mesh, area-weighted if $\rho \equiv 1$), and a point $s = (x, y) \in P$ with coordinates $x, y$ and density $\rho(s)$ there, define the second moments about the mass centroid $G = (G_x, G_y)$:

$$I_{xx} = \int_P \rho(s)\,(y - G_y)^2\,ds, \qquad I_{yy} = \int_P \rho(s)\,(x - G_x)^2\,ds, \qquad I_{xy} = \int_P \rho(s)\,(x - G_x)(y - G_y)\,ds$$

`moment_of_inertia_index()` already computes $I_{xx}$/$I_{yy}$ (their sum is $J$, its own polar moment) and, as of this index, $I_{xy}$ too - via the standard closed-form polygon product-of-inertia formula, the same triangle-by-triangle shoelace approach $I_{xx}$/$I_{yy}$ already use. The $2\times2$ symmetric matrix

$$M = \begin{pmatrix} I_{xx} & I_{xy} \\ I_{xy} & I_{yy} \end{pmatrix}$$

is the shape's own mass inertia tensor. Its eigenvalues $\lambda_{\min} \le \lambda_{\max}$ are the two **principal moments** - the second moments the shape would have about its own two natural (perpendicular) axes, whatever those happen to be, found without ever needing to know the axes' actual orientation. `moment_isotropy_index()` is $\lambda_{\min}/\lambda_{\max}$.

## Picturing the tensor: the equivalent ellipse

$M$ has an exact geometric picture - the **equivalent ellipse**, the unique ellipse with the same second moments as the shape. Its axes point along the tensor's eigenvectors and its semi-axis lengths are proportional to $\sqrt{\lambda}$, so it stretches the same way the shape's mass does: nearly circular for an isotropic shape, long and thin for an elongated one. The index is exactly the *squared* ratio of its minor to major axis.

```{r equivalent-ellipse, fig.width=9, fig.height=4}
# each shape gets its equivalent ellipse + principal axes, translated to a
# common local origin and then offset horizontally so the two sit side by
# side in one equal-aspect panel (facets can't combine free scales with
# coord_equal)
ellipse_of <- function(g, dx, k = sqrt(3), n = 120) {
  r <- moment_of_inertia_index(st_sfc(g))
  G <- st_coordinates(r$centroid)[1, 1:2]
  C <- matrix(c(r$Iyy, r$Ixy, r$Ixy, r$Ixx), 2, 2) / r$area  # (x,y) covariance
  e <- eigen(C, symmetric = TRUE)
  axes <- e$vectors %*% diag(k * sqrt(e$values))             # columns = principal semi-axes
  shift <- function(m) cbind(m[, 1] - G[1] + dx, m[, 2] - G[2])
  tt   <- seq(0, 2 * pi, length.out = n)
  ell  <- shift(t(G + axes %*% rbind(cos(tt), sin(tt))))
  ring <- shift(st_coordinates(st_sfc(g))[, 1:2])
  seg  <- data.frame(x = dx, y = 0,
                     xend = dx + c(axes[1, 1], axes[1, 2]),
                     yend = c(axes[2, 1], axes[2, 2]))
  lab  <- sprintf("isotropy %.3f", moment_isotropy_index(st_sfc(g))$index)
  list(ell = data.frame(x = ell[, 1], y = ell[, 2]),
       ring = data.frame(x = ring[, 1], y = ring[, 2]),
       axes = seg, label = data.frame(x = dx, y = 8.5, lab = lab))
}
demo   <- list(square = square, rect = rectangles[["aspect 4x"]])
offs   <- c(0, 26)
panels <- Map(ellipse_of, demo, offs)
grab <- function(k) do.call(rbind, lapply(names(panels), function(nm) {
  d <- panels[[nm]][[k]]; d$shape <- nm; d
}))

ggplot() +
  geom_polygon(data = grab("ring"), aes(x, y, group = shape),
               fill = "grey90", color = "grey40") +
  geom_path(data = grab("ell"), aes(x, y, group = shape),
            color = "firebrick", linewidth = 0.9) +
  geom_segment(data = grab("axes"), aes(x, y, xend = xend, yend = yend),
               color = "firebrick", linewidth = 0.5, linetype = "dashed") +
  geom_text(data = grab("label"), aes(x, y, label = lab), size = 3.5) +
  coord_equal() +
  theme_gallery
```

The square's equivalent ellipse is a circle (both principal moments equal, isotropy 1); the rectangle's is stretched to the same 4:1 ratio as the shape, so its minor-to-major axis ratio squared - $(1/4)^2 = 0.0625$ - is the index. Nothing about this needs the shape to *be* an ellipse; the equivalent ellipse only has to share its two second moments.

## Why this is always in (0, 1] - no rearrangement inequality needed

$M$ is provably positive semi-definite for *any* mass distribution, by construction rather than by any shape-specific argument: it's an integral of rank-1 PSD outer products, $M = \int_P \rho(s) \begin{pmatrix} y - G_y \\ x - G_x\end{pmatrix}\begin{pmatrix} y - G_y & x - G_x\end{pmatrix} ds$, where $s = (x, y)$. A sum (or integral) of PSD matrices is PSD, so $\lambda_{\min} \ge 0$ always, and $\lambda_{\max} > 0$ for any polygon with positive area - $\lambda_{\min} = 0$ only in the degenerate limit where every point of positive mass lies on a single line, never a genuine 2D polygon. That's the entire proof: no bathtub principle, no rearrangement inequality, no reference shape to derive - just a fact about the algebraic form of $M$ itself.

```{r rotation-invariance, echo=TRUE}
rect10 <- make_rect(sqrt(2000), sqrt(20))  # a 10x-aspect rectangle
res_unrotated <- moment_isotropy_index(st_sfc(rect10))

theta <- 0.37
rot <- matrix(c(cos(theta), sin(theta), -sin(theta), cos(theta)), 2, 2)
rect10_rotated <- st_polygon(list(st_coordinates(rect10)[, 1:2] %*% t(rot)))
res_rotated <- moment_isotropy_index(st_sfc(rect10_rotated))

data.frame(
  quantity  = c("Ixx", "Iyy", "Ixy", "index"),
  unrotated = c(res_unrotated$Ixx, res_unrotated$Iyy, res_unrotated$Ixy, res_unrotated$index),
  rotated   = c(res_rotated$Ixx, res_rotated$Iyy, res_rotated$Ixy, res_rotated$index)
) |> knitr::kable(format = "html", digits = c(0, 4, 4, 4), row.names = FALSE, escape = FALSE,
                   col.names = c("quantity",
                                 paste0(shape_thumb(rect10), "<br>unrotated"),
                                 paste0(shape_thumb(rect10_rotated), "<br>rotated 0.37 rad")))
```

$I_{xx}$, $I_{yy}$, and $I_{xy}$ all change under rotation - none of the three is rotation-invariant on its own. But $M$'s *eigenvalues*, and so the index built from them, don't - the difference between the two `index` values above is `r sprintf("%.2e", abs(res_unrotated$index - res_rotated$index))`, floating-point noise rather than any real movement. That's expected: $\lambda_{\min}$/$\lambda_{\max}$ are properties of the tensor itself, not of whatever coordinate axes happened to be used to write it down.

## Not "= 1 iff a disk"

```{r isotropic-shapes, echo=TRUE}
data.frame(
  shape = c(shape_thumb(disk), shape_thumb(square), shape_thumb(hexagon)),
  name = c("disk", "square", "regular hexagon"),
  isotropy = c(moment_isotropy_index(st_sfc(disk))$index,
               moment_isotropy_index(st_sfc(square))$index,
               moment_isotropy_index(st_sfc(hexagon))$index)
) |> knitr::kable(format = "html", digits = 8, row.names = FALSE, escape = FALSE)
```

All three score (numerically) exactly 1, not just approximately - a square and a hexagon have no preferred axis for their own mass to spread along, so their principal moments are exactly equal, same as a disk's. Contrast `moment_of_inertia_index()`, where only the disk reaches the ceiling and a square scores `r sprintf("%.3f", moment_of_inertia_index(st_sfc(square))$index)`: that index measures dispersal relative to a disk specifically, while isotropy measures anisotropy relative to *the shape's own two axes*, whatever they are. A shape can be maximally isotropy-1 while still being far from round - a hexagon isn't a disk, it just doesn't prefer a direction. The next two sections push on exactly how far that separation goes.

# Algorithmic choices

There isn't a `deterministic` argument, a quadrature parameter, or a Monte Carlo mode to discuss here, and it's worth being explicit about why rather than leaving the absence unexplained. $I_{xx}$, $I_{yy}$, and $I_{xy}$ are exact closed-form polygon integrals, summed triangle by triangle via the shoelace formula - there's no approximation layer for a `deterministic`-style argument to switch between. More than that, the sum is provably the same regardless of *which* valid triangulation produced the triangles: moments are additive over any partition of the polygon into non-overlapping pieces, so any two triangulations of the same region must sum to the same total, as long as each triangle's own contribution is computed with consistent (CCW) winding - which `.moment_of_inertia_core()`'s own winding check guarantees regardless of what a particular triangulator happens to produce.

```{r triangulation-invariance, echo=TRUE}
# a fan triangulation from every vertex of a convex, irregular heptagon -
# seven genuinely different triangulations of the exact same region
ang <- c(10, 50, 100, 160, 210, 270, 320) * pi / 180
r   <- c(5, 4.2, 5.5, 4.8, 5.1, 4.5, 5.3)
v   <- cbind(r * cos(ang), r * sin(ang))
heptagon <- st_polygon(list(rbind(v, v[1, ])))
n <- nrow(v)

fan_from <- function(apex) {
  ord  <- ((apex + 0:(n - 2)) %% n) + 1   # the other n-1 vertices, in cyclic order
  tris <- lapply(seq_len(n - 2), function(k) {
    st_polygon(list(rbind(v[apex, ], v[ord[k], ], v[ord[k + 1], ], v[apex, ])))
  })
  st_sf(area = vapply(tris, function(t) as.numeric(st_area(st_sfc(t))), numeric(1)),
        geometry = st_sfc(tris))
}

idx_cdt <- moment_isotropy_index(st_sfc(heptagon),
                                  prep = list(poly = heptagon, tri = cdt_triangles(st_sfc(heptagon))))$index
idx_fan <- vapply(1:n, function(apex) {
  moment_isotropy_index(st_sfc(heptagon), prep = list(poly = heptagon, tri = fan_from(apex)))$index
}, numeric(1))

data.frame(triangulation = c("CDT", paste("fan from vertex", 1:n)),
           index = c(idx_cdt, idx_fan)) |> knitr::kable(format = "html", digits = 10, row.names = FALSE)
```

All eight triangulations - the package's own CDT plus a fan from every one of the heptagon's seven vertices - agree to floating-point noise (max difference `r sprintf("%.1e", max(abs(idx_fan - idx_cdt)))`). The same holds on non-convex polygons via a diagonal flip of an existing CDT mesh (not shown here); it isn't specific to convex shapes. This is a genuinely different situation from `convexity_index()`/`span_index()`, both of which *do* depend on mesh resolution (finer quadrature or subdivision moves their value, even if only slightly) because they estimate an integral with no closed form. Moment isotropy has nothing to estimate.

# Illustrations

## Basic shapes: isotropy vs. the two properties it isn't

```{r basic-table}
rows <- c(list(square = square, hexagon = hexagon, disk = disk), rectangles, stars_r)
tbl <- do.call(rbind, lapply(names(rows), function(nm) {
  poly <- st_sfc(rows[[nm]])
  data.frame(shape = shape_thumb(rows[[nm]]),
             name = nm,
             isotropy = moment_isotropy_index(poly)$index,
             moment_of_inertia = moment_of_inertia_index(poly)$index,
             convexity = convexity_index(poly)$index)
}))
knitr::kable(format = "html", tbl, digits = 3, row.names = FALSE, escape = FALSE)
```

Elongating a rectangle (constant area, growing aspect ratio) collapses isotropy toward 0 fast - by the closed form for a $W \times H$ rectangle, $\lambda_{\min}/\lambda_{\max} = \min(H^2, W^2)/\max(H^2, W^2)$, so a 20x aspect ratio gives an index of exactly $1/400 = 0.0025$, matching the table above. `moment_of_inertia` falls too, but more gently - it's penalising dispersal from the centroid generally, not elongation specifically, so a stretched-but-not-dispersed shape doesn't hit its floor as hard.

The star sequence tells the opposite story: deepening notches drag `convexity` down sharply (that's what it's built to detect) while `isotropy` barely moves - a 6-pointed star with deep notches is still, on average, pulling mass roughly equally in six directions, so its principal moments stay close to equal even as it becomes badly non-convex. **Elongation and concavity are independent properties**, and this table is the same point the classic-metrics vignette makes with `moment_of_inertia_index()` vs `convexity_index()`, made again here with a third, independent axis.

## Weighting: isotropy is "free" with no separate reference to derive

Weighting is exactly as cheap here as for `moment_of_inertia_index()`, and for the same reason: $I_{xx}$/$I_{yy}$/$I_{xy}$ are already $\rho$-weighted sums over triangles, so a weighted isotropy is just the same eigenvalue ratio computed from the weighted tensor - there's no concentric-rings derivation to redo, because there's no reference shape in the first place.

```{r weighted-disk, echo=TRUE, fig.width=8, fig.height=4}
prep <- prepare_polygon(st_sfc(disk))
tri <- prep$tri
ang <- atan2(st_coordinates(st_centroid(st_geometry(tri)))[, 2],
             st_coordinates(st_centroid(st_geometry(tri)))[, 1])
w_ew <- 1 + 3 * cos(ang)^2   # weight concentrated along the east-west axis

plain <- moment_isotropy_index(st_sfc(disk), prep = prep)
weighted <- moment_isotropy_index(st_sfc(disk), prep = prep, weight = w_ew)

# the shape itself never changes here, so the thumbnail shows the WEIGHT
# distribution (shaded by density) rather than a fixed outline - what
# varies row to row is where the mass sits, not the boundary
data.frame(
  weighting = c(weight_thumb(tri, rep(1, nrow(tri))), weight_thumb(tri, w_ew)),
  name = c("uniform (area)", "concentrated east-west"),
  isotropy = c(plain$index, weighted$index)
) |> knitr::kable(format = "html", digits = 3, row.names = FALSE, escape = FALSE)
```

The disk's own boundary never changes - only how its mass is distributed within it - yet concentrating weight along one axis drops isotropy well below 1. A perfectly round *shape* is no guarantee of a round *mass distribution* once weighting enters the picture, the same lesson `vignette("f-understanding-radial-concentration-index")`'s "weight moves the reference" section makes for that index's own reference.

## Things to watch out for: isotropic tensor, not isotropic coverage

```{r pinwheel-table}
pinwheel_tbl <- do.call(rbind, lapply(names(pinwheels), function(nm) {
  poly <- st_sfc(pinwheels[[nm]])
  data.frame(shape = shape_thumb(pinwheels[[nm]]),
             name = nm,
             moment_isotropy = moment_isotropy_index(poly)$index,
             directional_balance = directional_balance_index(poly)$index,
             convexity = convexity_index(poly)$index)
}))
pinwheel_tbl <- rbind(pinwheel_tbl,
                       data.frame(shape = shape_thumb(disk),
                                  name = "disk (reference)",
                                  moment_isotropy = moment_isotropy_index(st_sfc(disk))$index,
                                  directional_balance = directional_balance_index(st_sfc(disk))$index,
                                  convexity = convexity_index(st_sfc(disk))$index))
knitr::kable(format = "html", pinwheel_tbl, digits = 4, row.names = FALSE, escape = FALSE)
```

From 3 arms up, `moment_isotropy` reads *exactly* 1 - identical, to the precision shown, to a genuine solid disk - despite wide empty sectors between the arms where there's no mass at all. This isn't a numerical coincidence: any rank-2 tensor built from vectors arranged with three-fold or higher rotational symmetry is forced to be isotropic, a fact from representation theory rather than anything specific to polygons or mass distributions. `directional_balance_index()` is fooled the same way, for a related but distinct reason - see `vignette("h-understanding-directional-balance-index")`'s own "things to watch out for" section, where a 2-armed (dumbbell) shape cancels out because that index only needs *two-fold* symmetry to zero its first moment. Here, the second moment needs three-fold or higher - which is exactly why the 2-armed case above is the one row where `moment_isotropy` correctly reads low. Only `convexity_index()` catches any of these pinwheels for what they visually are.

The distinction worth carrying forward: `moment_isotropy_index()` measures whether a shape's mass is spread *equally* across directions - the two principal moments come out equal - not whether mass is *present* in every direction. A shape built entirely from a handful of narrow, evenly-spaced spokes satisfies the first condition perfectly while failing the second about as badly as a shape can. Reach for `convexity_index()`, not this index, when the question is really "does this shape have gaps."

# Where this fits

`moment_isotropy_index()` is a genuinely different question from this package's other eleven indices, not a fifth entry in a family: "is this shape convex" (`convexity_index()`, `hull_ratio_index()`), "how far is mass dispersed from the centre" (`moment_of_inertia_index()`, `span_index()`, `radial_concentration_index()`), "how round is the boundary/bounding shape" (`polsby_popper_index()`, `reock_index()`, `width_length_ratio_index()`), "does the mass distribution prefer an axis" (`moment_isotropy_index()`), and "does the mass distribution prefer a bearing" (`directional_balance_index()`, see `vignette("h-understanding-directional-balance-index")`) are five separate properties that happen to correlate on many real shapes without being redundant with each other. Two direct side-by-side comparisons already sit above rather than being repeated here: the basic-shapes table in "Illustrations" (isotropy vs. `moment_of_inertia_index`/`convexity_index` on the same shapes) and the pinwheel table in "Things to watch out for" (isotropy vs. `directional_balance_index`/`convexity_index` on shapes designed to expose each index's own blind spot).

# Key takeaways

- `moment_isotropy_index()` measures whether a shape's mass is spread *equally* across directions (its two principal moments come out equal), not whether mass is *present* in every direction - a handful of narrow, evenly-spaced spokes with wide empty gaps between them can satisfy the first condition perfectly while failing the second badly.
- It is exact and closed-form throughout: no `deterministic` argument, no quadrature, no Monte Carlo mode, and provably triangulation-invariant (confirmed above across eight different valid triangulations of the same polygon, agreeing to floating-point noise) - unlike `convexity_index()`/`span_index()`, there is no approximation layer for a mesh choice to move.
- It's bounded in $(0, 1]$ by an elementary matrix fact (the mass inertia tensor is always positive semi-definite), not a rearrangement inequality or bathtub-principle argument - the only index in this package's family with that kind of proof.
- An index of 1 does not mean "this is a disk": any shape with three-fold or higher rotational symmetry - a square, a hexagon, or the pinwheels above - reaches exactly 1, whatever its actual boundary looks like. Reach for `convexity_index()`, not this index, when the real question is "does this shape have gaps."
- Weighting is free: since there's no reference shape to re-derive (unlike `moment_of_inertia_index()`'s concentric rings or `span_index()`'s annulus), a weighted isotropy is just the same eigenvalue ratio computed from the reweighted tensor.
- It rides loosely with the "dispersal from a centre" family (`moment_of_inertia_index()`, `span_index()`, `radial_concentration_index()`) on real shapes, but is structurally the odd one out - the only member immune to *where* the mass's centre sits, only to the *shape* of the spread around it.
